---
title: "My Daily Productivity Dataset"
author: "N. O. Body"
date: "`r format(Sys.Date(), '%B %Y')`"
output: pdf_document
---

## Introduction


Over a period of thirty consecutive days, I recorded several aspects
of my daily routine to explore how coffee consumption and sleep might
influence my productivity. Each evening, I noted how many cups of
coffee I had consumed before noon, how many hours I had slept the
previous night, my self-rated morning mood, whether I considered the
day productive, and how many tasks I had completed. This dataset
reflects genuine features of my everyday life and captures factors
that I believe affect my ability to work effectively. Understanding
these relationships is personally meaningful because productivity,
sleep, and caffeine intake are recurring themes in my daily routine.

Below are the first few rows of the dataset:

```{r setup, include=TRUE, message=FALSE}
# Read the dataset
df <- read.table("dataset.txt", header = TRUE, sep = "\t")

# Display first few rows
head(df)
```


Before applying statistical tests, I explored the data visually to understand its structure.


```{r}
# Histogram of tasks completed
hist(df$tasks_completed, 
     main = "Distribution of Tasks Completed", 
     xlab = "Tasks Completed", 
     col = "lightblue", border = "white")

# Scatterplot of coffee vs tasks
plot(df$cups_coffee, df$tasks_completed,
     pch = 19, col = "darkblue",
     xlab = "Cups of Coffee", 
     ylab = "Tasks Completed",
     main = "Tasks vs Coffee Consumption")
abline(lm(tasks_completed ~ cups_coffee, data = df), col = "black", lwd = 2)
```

## Student's t-test

I used a t-test to compare the mean number of tasks completed on days
I considered "productive" versus "non-productive".

Null hypothesis ($H_0$): The mean number of tasks completed is the
same on productive and non-productive days.  Alternative hypothesis
($H_1$): The mean number of tasks completed differs between productive
and non-productive days.

```{r}
t_test_result <- t.test(tasks_completed ~ productive_day, data = df)
t_test_result
```

Interpretation: The p-value indicates whether there is a statistically
significant difference in average task completion between the two
types of days.  If $p < 0.05$, I reject $H_0$, suggesting that my
perception of productivity corresponds to a real difference in work
output.


# Linear regression

Next, I examined whether coffee consumption predicts the number of
tasks completed.

Model: tasks_completed ~ cups_coffee

$H_0$: The slope = 0 (no linear relationship).\\
$H_1$: The slope not = 0.


```{r}
fit <- lm(tasks_completed ~ cups_coffee, data = df)
summary(fit)

# Extract slope p-value and labels
p_value <- summary(fit)$coefficients[2, 4]
sig_label <- ifelse(p_value < 0.05,
                    paste0("Slope significant (p = ", signif(p_value, 3), ")"),
                    paste0("Slope not significant (p = ", signif(p_value, 3), ")"))
eq_label <- paste0("Tasks = ", round(coef(fit)[1], 2), " + ",
                   round(coef(fit)[2], 2), " x Coffee")

# Plot
plot(df$cups_coffee, df$tasks_completed,
     pch = 19, col = "blue",
     xlab = "Cups of Coffee",
     ylab = "Tasks Completed",
     main = "Linear Regression: Tasks vs Coffee")
abline(fit, col = "black", lwd = 2)
usr <- par("usr")
text(x = usr[1] + 0.5, y = usr[4] - 0.5, labels = eq_label, adj = c(0,1), cex = 0.9)
text(x = usr[2] - 0.5, y = usr[3] + 0.5, labels = sig_label, adj = c(1,0), cex = 0.9)
```

Interpretation: The regression output shows how each additional cup of
coffee is associated with an increase in the number of tasks
completed.  The p-value for the slope quantifies whether this
association is statistically significant.


## Fisher's Exact Test

To investigate whether high coffee consumption is associated with
perceived productivity, I created a binary variable indicating whether
I drank three or more cups of coffee.

$H_0$: Coffee consumption level and productivity perception are independent.\\
$H_1$: Coffee consumption level and productivity perception are associated.


```{r}
df$high_coffee <- ifelse(df$cups_coffee >= 3, "High", "Low")
table_summary <- table(df$high_coffee, df$productive_day)
table_summary
fisher_result <- fisher.test(table_summary)
fisher_result
```



# Conclusions

This small self-collected dataset offered an engaging way to apply
statistical methods to my own life. The t-test confirmed that I
genuinely complete more tasks on days I perceive as productive. The
linear regression indicated a positive relationship between coffee
intake and task completion, although the residual variation suggests
that other factors---such as sleep or workload---also play an important
role. The Fisher's exact test suggested a modest association between
high coffee intake and productivity perception.

Collecting and analysing my own data made statistical reasoning feel
concrete and relevant. I was surprised that the coffee-productivity
link, while statistically significant, was not as strong as I had
expected. This exercise reinforced the value of evidence-based
thinking, even in the small decisions of daily life.
