# Getting started {#introduction}


## The R programming language

The R programming language is used extensively in this course.  The
final exam and the midsemester test will require you to use R to
perform statistical analyses.  R is free software, which means that
you can download and install it at zero cost.  This also means that
you are free to modify that code as you wish, although this will not
be part of stat500.

You can download R from the CRAN website:

```
https://cran.r-project.org/
```

(simply click on "download" and follow the instructions).  Many people
prefer to use Rstudio which makes R easier to use, and allows students
to use the course material more efficiently.  It too is free, and
available at

```
https://www.rstudio.com/
```

Both R and Rstudio have versions for Windows, Mac, and Linux.

R is the de-facto standard for statistical analysis and it is required
for many, many jobs in industry, commerce, science, and engineering.


## Statistics and R

Statistics is the art of learning from data.  It is concerned with the
collection of data, it subsequent description and analysis, which
often leads to the drawing of conclusions.

There are two main branches of statistics:

* _Descriptive Statistics_: describing and summarizing of data
* _Inferential Statistics_: drawing of conclusions from data

The first part of STAT500 will discuss descriptive statistics and the
second will discuss inferential statistics.


## Data

Although we think we know what "data" means, it is difficult to pin
down the term without using circular definitions.  Here we will
content ourselves with giving some examples of data:

* Bus arrival times at Britomart
* Number of packets of M\&Ms sold daily for a month
* Length of time customers queue at Disneyland's Space Mountain
* Amount of money spent on advertising by different organisations
* Grams of protein per litre in milk from different locations
* Amount of rainfall in Auckland over past 100 years
* Possession and territory in a rugby match

Data may be collected from existing sources of information, it may be
observed or it may be collected intentionally (e.g. through an
experiment).  Here we are not going to focus on the collection of
data.

## Samples and populations

In statistics, _population_ refers to the total set of observations
that could possibly be made.  For example, if we are studying the
weight of adult New Zealanders, ythen the population is the set of
weights of all New Zealanders.  If we are studying sand grains, then
the population is all sand grains in the world.

A _sample_ is a group of representative individuals from the
population.  We use the sample to make *inferences* about the
population.  Often the data we analyse will be from a sample rather
than a population.

Technology makes it easier to take larger samples, but a _census_ (a
survey of the entire population) is often impractical or impossible,
so we are forced to consider samples.  Samples are generally easier to
deal with and manage than populations.  Quite often we use the concept
of _random sample_ in statistics.  A random sample has the feature
that every member of the population has the same probability of being
chosen.

When it comes to samples, size matters.  How data was collected
influences techniques that can be used to analyse the data.  Even when
using data collected by someone else (secondary data), we need to be
aware how the data was collected.


We will use R for some descriptive statistics below.

## Getting started in R

At its simplest, R can be used as a calculator.  Fire up R or Rstudio,
click in the console, and type some math:

```{R}
2+2
17*33
6/19
```

In the above, the R commands you type at the command prompt are shown
with a grey background, and R's response appears next, preceded by two
hash signs (different implementations have slightly different
appearances).  You are strongly encouraged to follow the presentation
in this course manual, typing the commands as you go.  This is the
only way to learn!

You can also define variables and then use that variable in another
calculation

```{r}
x <- 19
x
```

In the above, the variable `x` is assigned the value 19, and on
the next line we print x just to check.

```{r}
y <- x+5
y 
```

There are some in-built functions (e.g. `mean()`) that can be used
in your computations.

```{r}
v <- c(2,5,4,3,4,3,2,3,2,3)
mean(v)
```

## Data visualization and summary with R

R can be used to prepare visual and numerical summaries of datasets.
R includes a large number of builtin datasets, and we will use the
`cars` dataset first.

```{r}
data(mtcars)
mtcars
```

In the above, the dataset `mtcars` is a dataset of cars taken from
the 1974 Motor Trend magazine; to get more information, type
`?mtcars` at the R command prompt which will fire up R's help
system.  See how each row corresponds to a car and each column
corresponds to a property of that car.  We can extract a single column
with the dollar operator:

```{r}
mtcars$mpg
```

and we can see the mpg [miles per gallon; a measure of fuel
efficiency] for each car; but it is difficult to get a sense of the
dataset from this output format.

## Summary statistics

It is often helpful to summarize a dataset using a single number that
represents all the observations.  There are many ways to do this but
the three most common are *mean*, *median* and *mode*.

### Mean

We have the mpg rating for each car but we can summarize the whole
dataset by taking the *mean*.  Mathematically, the mean of a dataset
is the sum of all the observations, divided by the number of
observations.

If we have observations $x_1,x_2,\ldots, x_n$ [that is, the first
observation is $x_1$, the second observation is $x_2$, and so on, up
to $x_n$ being the $n^\mathrm{th}$ observation], then we write
$\overline{x}$ for the mean and calculate it as follows:

\[
\overline{x} = \frac{x_1+x_2+\ldots +x_n}{n}
\]

But sometimes we use the summation notation:

\[
\overline{x} = \frac{1}{n}\sum_{i=1}^n x_i
\]

It is easy to calculate the mean in R using the `mean()` function:


```{r}
mean(mtcars$mpg)
```

### Median

The _median_ is the middle point of the data: half the datapoints are
larger than the median and half are smaller.  Calculating this is easy
with R function `median()`:

```{r}
median(mtcars$mpg)
```

### Mode

The _mode_ of the data is the observation that occurs most frequently.
This can be problematic sometimes but in the case of the cars dataset
we can work with the number of cylinders:

```{r}
mtcars$cyl
```

It's not clear from the above which is the most common number of
cylinders, but we can summarise the dataset using `table()`:

```{r}
table(mtcars$cyl)
```

from which we see that the mode is 8 cylinders, as 14 cars have this
number of cylinders, more than any other number.


## Spread

We are often interested in the spread of data: what is the range of
the datapoints? How widely are the observations distributed?  Are they
all confined to a very narrow position, or are they widely scattered?

One simple answer to this is to state the *range*:

```{r}
x <- mtcars$mpg
min(x)
max(x)
```

(In the above, I have used a variable `x` to save typing).  Thus
the range is from 10.4 to 33.09.

Other measures of spread are the _standard deviation_ which has a
rather technical definition but is understandable as an average
distance from the mean (we will discuss this when we learn about the
Gaussian distribution).  It is easy to calculate in R with the
`sd()` command:

```{r}
sd(x)
```

Other measures are the interquartile range, which gives the range
within which half the data lies.  This is most easily calculated
using the `summary()` command, which gives other information too:

```{r}
summary(x)
```


## Visual analysis

While measures such as the mean and standard deviation are
informative, presenting data visually is often a powerful medium for
understanding and communicating datasets.  Visual analysis is made
easy in R with a very wide range of functionality.  Here we will
discuss a few techniques.  We can produce a histogram using the
`hist()` function:

```{r}
hist(mtcars$mpg)
```

This produces a rather plain histogram.  R provides ways to make the
histogram more colourful, and we can use constructions such as
`hist(mtcars$mpg,col="red")`.  See the help page (`?hist`) for
details.  We can produce a slightly more informative, but simpler,
presentation, using the `stem()` function:


```{r}
stem(mtcars$mpg)
```


Distributions can have different shapes, and one way to quantify this
is to calculate the *skew*:

```{r, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
par(mfrow=c(1,3))
jj <- qchisq(0.5,df=3)
x <- seq(from=0,to=10,len=500)
plot(x,dchisq(x,df=3),lwd=5,type='l',axes=F,xlab='',ylab='',main='positive skew')
abline(h=0, v=c(1,jj,3))
text(1,0.06,'mode',pos=4,offset=0)
text(jj,0.04,'median',pos=4,offset=0)
text(3,0.02,'mean',pos=4,offset=0)
xx <- seq(from=-3,to=3,len=500)
plot(xx,dnorm(xx),lwd=5,type='l',axes=FALSE,xlab='',ylab='',main='zero skew')
abline(h=0,v=0)
text(0,0.06,'mode',pos=4)
text(0,0.04,'mean',pos=4)
text(0,0.02,'median',pos=4)
plot(x,dchisq(10-x,df=3),lwd=5,type='l',axes=FALSE,xlab='',ylab='',main='negative skew')
abline(h=0, v=10-c(1,jj,3))
text(9,0.06    ,'mode'   ,pos=2,offset=0)
text(10-jj,0.04,'median' ,pos=2,offset=0)
text(7,0.02    ,'mean'   ,pos=2,offset=0)
```

In the above we see a positively-skewed distribution, a symmetrical
distribution, and a negatively-skewed distribution.


## Bivariate analysis

It is also possible to plot one variable as a function of another, in
terms of a scattergraph:

```{r}
plot(disp~hp,data=mtcars)
```

In the above, the R command plots `disp` (displacement, the volume
of the car engine) on the vertical axis against `hp` (horsepower,
a measure of the engine power) on the x axis.  The tilde symbol
("`~`") is read "is a function of"; thus the command reads "plot
`disp`, as a function of `hp`, for dataset `mtcars`".
Also Note that we need to specify the datasource with the `data=`
argument to the `plot()` command.  Later in the course we will see
ways in which we can analyse such datasets more deeply.

Boxplots are produced using the `boxplot()` command:

```{r}
boxplot(hp~cyl,data=mtcars)
```

The above plot shows how hp varies with `cyl`, the number of
cylinders in the car.  We will discuss the boxplot in more detail
later in the course, but for now see how cars with more cylinders
produce more horsepower.

