# Probability


Probability and randomness are very difficult to define without using
circular logic, but easy to recognise.

In loose terms, probability is a measure of one's confidence that an
event will occur.

*  What is the probability that it will rain tomorrow?
*  What is the probability that I will pass my statistics exam?
*  What is the probability that the train will be late?
*  What is the probability that a particular stock price will increase?
*  What is the probability that I will roll 3 sixes in a row?

Some of the questions above have well-defined and objective answers,
some do not.  Probability and randomness are subtle concepts and there
are different interpretations of probability statements.

In this course, we will interpret probability as the long-run relative
frequency of an event occurring.  For example, suppose a fair coin is
flipped, then the probability of getting a head can be interpreted as
the relative frequency of getting a heads over a very large number of
trials.  This will be close to one-half if the coin is truly fair.

(An alternative interpretation is subjective probability.  This is a
measure of an individual person's belief that an event will occur. We
will not consider subjective probability in this paper, but it is used
in more advanced courses and is becoming the dominant methodology in
certain scientific areas including climate change and astrophysics).

To start thinking about probability, we need some basic
terminology.

## Random variables

A *variable* is a characteristic that changes or varies over time
and/or for different individuals or objects under consideration

Examples:

* In a survey of university students, height, gender, income and age will all vary from person to person.
* The number obtained when rolling a die will vary with each roll.
* The price of ANZ shares will vary over time.
* whether the All blacks in or not in a game will change from match to match

A *random variable* (usually denoted with a capital letter, usually
$X$) is a variable whose value is determined by the measuring the
outcome of a random process.  For example:

* $X$ = Number of defects on a randomly selected piece of furniture
* $X$ = Number of girls in a randomly selected family with 3 children
* $X$ = Number obtained when rolling a die
* $X$ = The height of a randomly selected AUT student
* $X$ = Time taken to respond to a randomly selected call to 111

(the above definition is somewhat circular---it uses randomness to
define random---but giving a watertight definition requires several
years of postgraduate study in pure mathematics even to state.  We
will use terminology somewhat loosely in this course).

## The language of probability

An *experiment* is any process that produces an observation or
outcome.  The set of all possible outcomes of an experiment is called
the *sample space*.

If the experiment is tossing a coin, the sample space will be the set
$\{\mathrm{head}, \mathrm{tail}\}$.  If it is throwing two dice, the
sample space will ordered pairs $(a,b)$ where $a,b$ are members of the
set $\{1,2,3,4,5,6\}$.  Continuous variables are more problematic.  We
usually take the sample space to be the real numbers, or sometimes the
computer representation of the real numbers.

An *event* is a set of outcomes of the experiment, or a subset of the
sample space.  For example, if the experiment is tossing a six-sided
die, then an event might be "obtaining an even number", that is, the
result being in the set $\{2,4,6\}$.  The simplest events are just
single points of the sample space.

### Example: three coloured balls in a bag

A box contains three balls—one red, one green, and one blue.  Consider
an experiment that consists of withdrawing a ball from the box, noting
its colour, replacing it, and then withdrawing a second ball.

Here the sample space will be, with obvious notation,

RR, RG, RB, GR, GG, GB, BR, BG, BB

It might be easier to see what is going on by arranging the elements
more logically:

RR, RG, RB

GR, GG, GB

BR, BG, BB


See how the elements of each row have the same first colour, and the
elements of each column have the same colour.  This makes it easier to
check that you have written the sample space out correctly.

## Set theory

We are going to need a little set theory; see figure \@ref(fig:venn)
which defines set intersection, set union, set complement and
symmetric difference.  

```{r venn, fig.cap='Two sets, $A$ and $B$ on a Venn diagram together with red areas marking set union ($A\\cup B$), set intersection ($A\\cap B$), complement of A ($\\overline{A}$) and symmetric difference ($A\\Delta B$)', echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics("stat601/intersection_union.pdf")
```

The *empty set* (written $\varnothing$) is the set with no elements.
Thus the set of billionaire statistics lecturers is the empty set (but
I'm working on it).

The *union* of $A$ and $B$ (written $A\cup B$) is the set of elements
that are in either $A$ or in $B$, or in both.  Sometimes we say that
the union is ``A or B''; but note that the English language is
somewhat imprecise here: if I say to my daughter that she can have an
ice cream or a drink, I mean that she has to choose one or the other.
She can't have both.

The *intersection* of $A$ and $B$ (written $A\cap B$) is the set of
elements that are in both $A$ and in $B$.  Thus if $A$ is the set of
all European cities, and $B$ is the set of all cities with a
population of more than 3 million, then $A\cap B$ is the set of all
European cities with a population of more than 3 million.  If two sets
have empty intersection (that is, $A\cap B=\varnothing$), we say that
$A$ and $B$ are *disjoint*.  Thus if $A$ is the set of billionaire
philanthropist playboy vigilantes, and $B$ is the set of AUT
statistics lecturers, then if a person is a member of set $A$ then he
will not be a member of set $B$; and conversely.  The sets are
disjoint^[Yup, definitely disjoint.  Absolutely empty intersection
there.  Yes sir.  No way these two sets could have any common members.
Ask my butler.].

The *complement* of event $A$ (written $\overline{A}$; sometimes in
books you will see $A'$ or $A^c$), is the set consisting of all
elements which are not in $A$.  For example, if we are sampling people
in the street, we could define $A$ to be the event that the person is
a smoker.  Then the complement, $\overline{A}$ would be that the
person is not a smoker.

The *symmetric difference* of $A$ and $B$ (written $A\Delta B$) is
$\left(A\cup B\right)\cap\overline{A\cap B}$.  It is the set of all in
$A$ but not $B$, or in $B$ but not $A$.  This is what I mean when I
say to my daughter that she may have an ice cream or a drink.  I ought
to say to her: "you've been a good girl today, you can choose any
member of the set comprising the symmetric difference of the sets $A$
(the set of having an ice-cream) and $B$ (the set of having a drink)".
But note that this _obliges_ her to choose one or the other; she is
not at liberty to decline my offer.  She cannot have neither.

## Basic properties of probability

Here we will cover some basic facts about probability which are
intuitively obvious but might be phrased in unfamiliar wording.

### Probability of the sample space

If we have a set $A$ then the number of elements in $A$ is written
$n(A)$.  The *probability* of an event $A$ is denoted $p(A)$, or
sometimes $Pr(A)$.  If events corresponding to single elements of $S$
are equally likely, we _define_ the probability of $A$ to be
$\frac{n(A)}{n(S)}$.  We have that $p(A)$ is between zero and 1,
written $0\leq p(A)\leq 1$.  The probability of the sample space $S$
is $\frac{n(S)}{n(S)}=1$; we would write $p(S) = 1$.

### Probability of set union and the principle of inclusion-exclusion

It is one of the axioms of probability that $p(A\cup B)\leq
p(A)+p(B)$.  If two events $A$ and $B$ are disjoint, we have that
$p(A\cup B) = p(A) +p(B)$.  If $A$ and $B$ are not disjoint, the
situation is more complicated because we are double-counting elements
in $A\cap B$. We can also deduce that

\[
p(A\cup B)=p(A)+p(B)-p(A\cap B)
\]

which is known as the principle of inclusion-exclusion.  If we have
three sets the principle becomes more complicated still and we have

\[
p(A\cup B\cup C)=p(A)+p(B)+p(C)-p(A\cap B)-p(A\cap C)-p(B\cap C)+p(A\cap B\cap C)
\]


### probability of set complement

The *complement* of event $A$, denoted $\overline{A}$ (sometimes in
books you will see $A'$ or $A^c$), is the set consisting of all
outcomes in the sample space which are not in $A$.  For example, if we
are sampling people in the street, we could define $A$ to be the event
that the person is a smoker.  Then the complement, $\overline{A}$
would be that the person is not a smoker.  Because $A$ and
$\overline{A}$ are disjoint, we have that

\[
p(A) + p(\overline{A})=p(A\cup\overline{A})=p(S)=1.
\]

Therefore

\[
p(\overline{A}) = 1 - p(A)
\]

In words, this says "the probability of the complement of $A$ is one
minus the probability of $A$".  Thus if $p(A)=0.35$ then the
probability of being a non-smoker is 1-0.35, or 0,65.  This makes
sense intuitively.

## de Morgan's laws

de Morgan's laws are  the intuitively obvious set equations:

\[
\overline{A\cup B} = \overline{A}\cap\overline{B}
\]

and

\[
\overline{A\cap B} = \overline{A}\cup\overline{B}
\]


Take the first one as an example.  This says that the probability of
being "neither $A$ nor $B$" is equal to the probability of being "not
$A$" and "not $B$".


## Conditional probability

Quite often, two events are interdependent in the sense that one
conveys information about the other.  For example if $A$ is the event
"I have a glass of wine with my dinner" and $B$ is the event "my wife
Lesley has a glass of wine with her dinner", then knowing that $A$ has
occurred changes the probability that $B$ occurs (because if we open a
bottle of wine then we both have some).  We might have $p(A)=0.2$ and
$p(B)=0.1$.

But if Lesley has wine, then it is almost certain that I will have
some too.  We express this in probability language using the concepts
of conditional probability.  We might ask what the probability that I
have wine is, _given_ that Lesley has wine; this would be written
$p(A|B)$ (and in words is "the probability of $A$ given $B$").  We
might have $p(A|B)=0.9$.  So the probability of me having wine, given
that Lesley is having wine, is 90\%.

The conditional probability law states that

\[
p(A|B)=\frac{p(A\cap B)}{p(B)}
\]

(to verify this, look at figure \@ref(fig:venn)).

Quite often, events do not have any connection with one another, and
we have that $p(A) = p(A|B)$. In words, this means that the
probability of $A$ occurring is equal to the probability of $A$
occurring, given that $B$ has occurred.  If this is the case we say
that $A$ and $B$ are *independent*.  If two events are independent
then we have

\[
p(A) = p(A|B)=\frac{p(A\cap B)}{p(B)}
\]

so we can deduce that independent events satisfy $p(A\cap B)=p(A)p(B)$
which can serve as an alternative definition of independence.

## Example: rolling two dice

Consider an experiment which consists of rolling two dice and
observing the number on the uppermost face.  All of the outcomes of
this experiment form the *sample space*:

\begin{align*}
S = \{	& (1, 1), (1, 2), (1, 3), (1, 4), (1, 5), (1, 6), \\
	& (2, 1), (2, 2), (2, 3), (2, 4), (2, 5), (2, 6), \\
	& (3, 1), (3, 2), (3, 3), (3, 4), (3, 5), (3, 6), \\
	& (4, 1), (4, 2), (4, 3), (4, 4), (4, 5), (4, 6), \\
	& (5, 1), (5, 2), (5, 3), (5, 4), (5, 5), (5, 6), \\
	& (6, 1), (6, 2), (6, 3), (6, 4), (6, 5), (6, 6)\}
\end{align*}

In the above, note how the order of the numbers matter: we are
treating $(1,2)$ as a different result from $(2,1)$.   

Define events $A$ and $C$ as follows:

* event $A$: first number is 4 or 5
* event $B$: sum is 9 or greater (that is, if the throw is $(a,b)$ then $a+b\geq 9$)
* event $C$: first number is strictly less than or equal to the second number

Using the fact that $p(A) = n(A)/n(S)$ (and we know that $n(S)=6\times
6=36$), we can see that

* $A=\left\{(4,1),(4,2),(4,3),(4,4),(4,5),(4,6),(5,1),(5,2),(5,3),(5,4),(5,5),(5,6)\right\}$, so therefore $n(A) = 12$ and therefore $p(A)=n(A)/n(S)=12/36=1/3$.
* $B=\left\{(3,6),(4,5),(4,6),(5,4),(5,5),(5,6),(6,3),(6,4),(6,5),(6,6)\right\}$, so therefore $p(B) = 10/36=5/18$
* $C=\left\{(1,2),(1,3),(1,4),(1,5),(2,3),(2,4),(2,5),(2,6),(3,4),(3,5),(3,6),(4,5),(4,6),(5,6)\right\}$, so therefore $p(C) = 14/36=7/18$


We can find probabilities for the intersection and union easily too.
For example, if we wish to find $A\cap B$, then this is elements that
are in set $A$ and set $B$.  Thus we have to find throws that have the
first number 4 or 5, and also whose sum is $\geq 9$.  This would be
the set $\left\{(4,5),(4,6),(5,4),(5,5),(5,6)\right\}$, so $p(A\cap
B)=5/36$.

### Example question 1

Question: Find $p(A|B)$.

Answer: from the conditional probability formula, $p(A|B) =
\frac{p(A\cap B)}{p(B)} = \frac{5/36}{5/18}=\frac{1}{2}$.

### Example question 2


Question:  find $p(A\cup B)$.

Answer: From the principle of inclusion-exclusion, we have

\[
p(A\cup B)=p(A)+p(B)-p(A\cap B) = 
\frac{1}{3}+\frac{5}{18}-\frac{5}{36}=\frac{17}{36}
\]

### Example question 3

Question: Are events $A$ and $B$ independent?

Answer: independent events are characterised by $p(A\cap B)=p(A)p(B)$.

We have $p(A\cap B)=\frac{5}{36}$ but $p(A)p(B)=\frac{1}{3}\times\frac{5}{18}=\frac{5}{56}$ so the events are not independent.

### Example question 4

Question: Calculate $p(A\cup B\cup C)$.

Answer: exercise for the reader.


## Example: grades

Suppose we have a class of students who take a statistics course and a
physics course.  The pass rate for the statistics course is 60\% and
the pass rate for the physics course is 70\%.  We also know that 50\%
of students pass both physics and statistics.  This is shown in the
form of a Venn diagram in figure \@ref(fig:statisticsphysics).

```{r statisticsphysics,fig.cap="Pass rates for statistics and physics.  The areas of the regions are accurately presented", echo=FALSE,message=FALSE,warning=FALSE}
library("VennDiagram")
venn.plot <- draw.pairwise.venn(60, 70, 50, c("statistics", "physics"))
grid.draw(venn.plot)
```


### Example question 1

Question:  what is the probability of passing either physics or statistics or both?

Answer: we seek $p(S\cup P)$.  By the principle of
inclusion-exclusion, we have $p(S\cup P)=p(S)+p(S)=p(S\cap
P)=0.6+0.7-0.5=0.8$.  Students have an 80\% probability of passing at
least one paper.

### Example question 2

Question:  what is the probability of failing both?

Answer: we seek $p(S'\cap P')$.  by de Morgan, this is equal to
$1-p(S\cup P)$=1-0.8=0.2$.  Students have a 20\% probability of
failing both.

### Example question 3

Question:  A student is known to have passed physics.  What is the probability that they pass statisics?

Answer: We seek $p(S|P)$.  By the formula for conditional probability,
we have $p(S|P)=\frac{p(S\cap P)}{p(P)}=\frac{0.5}{0.7}=\frac{5}{7}$.
Thus a student who passes physics has a probability of $5/7$ of
passing statistics.


### Example question 3

Question:  Are the events "passing statistics" and "passing physics" independent?

Answer: If $S$ and $P$ are independent, we have $p(S\cap P)=p(S)\times
p(P)$.  We are told that $p(S\cap P)=0.5$, but $p(S)\times
p(P)=0.6\times 0.7=0.42$ which is different.  The events are not
dependent.



## Tree diagrams

Tree diagrams can be used to depict a series of events. They are
particularly useful when computing probabilities associated with a
series of events.

* Each node represents an event. 
* Branches are used to indicate connections between events.
* The probability of each event is written along the branch. 
* The probabilities on all branches stemming from the same event must sum to one.
* Joint probabilities can be found by multiplying along the branches and are often written at the end of a sequence of branches. 
* All of the joint probabilities should sum to one.


## Example of a tree diagram: testing for a disease

It is known that 0.5\% of the population have a particular disease.
Of those that have the disease, 99\% will test positive for the
disease.  Of those that do not have the disease, 2\% test positive for
the disease.  We can represent this information as a tree diagram, as
seen in figure \@ref(fig:tree).  This defines event $D$ as a randomly
chosen person has the disease, and Event $E$ as that person has a
positive test.

```{r tree, out.width="6in",fig.cap='A tree diagram showing diagnosis of a disease', echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics("tree_diagram.pdf")
```

Observe that this test is a "good" test: if you have the disease, the
test probably detects this; and if you do not have the disease, the
test probably says that you are clear.


### Example 1
Question: what is the probability that a person chosen at random has the disease?

Answer: from the data, $P(D) =0.005 = 0.5\%$

### Example 2

Question: what is the probability that a person chosen at random
received a positive test?

Answer: There are two possibilities for receiving a positive test.
Either you have the Disease and test positive, positive or you do not
have the disease and test positive (this is known as a "false positive").

These two possibilities are $D \cap E$ and $D' \cap E$, which are
disjoint.  Using the law of conditional probability we have $P(D \cap
E) = P(E|D) P(D) = 0.00495$ and $P(D' \cap E) = P(E|D') P(D') =
0.0199$.

Thus $p(E)=p(D\cap E)+p(D' \cap E) = 0.00495+0.0199=0.02485$, or a
little under $2.5\%$.

Note carefully that the majority of people with a positive test are
actually free of the disease.  Think about this and ask yourself what
medical implications such analysis might have.  Some diseases involve
painful and potentially lethal treatment, and the treatment is
pointless unless you actually have the disease.

### Example 3

Question: Are events D and E independent? Justify your answer.

Answer: $D$ and $E$ are independent if $P(E \cap D)=P(E)P(D)$.  We
have that $P(E \cap D) = 0.00495$ but $P(E)P(D) = 0.024855\times
0.005=0.00012$ approximately, so the events are not independent.
Observe that the test is useful *because* events $E$ and $D$ are not
independent.

## Probability distributions  (discrete)

(recall the definition of random variable given at the beginning of
this chapter).  Random variables may be discrete or continuous
depending on the type of values they may attain.  We will consider
discrete random variables first and deal with continuous random
variables later.  Suppose $X$ is a random variable that may take one
of $n$ values $x_1,x_2,\ldots,x_n$.  If the probability that $X$ takes
the value $x_1$ is $p_1$, we write $p(X=x_1)=p_1$.  Similarly,
$p(X=x_2)=p_2$, and so on up to $p(X=x_n)=p_n$.  Other notation might
be $p(x_1)$ for $p(X=x_1)$.

For example, we might ask a series of people which of red, green, or
blue is their favourite colour.  The random variable $X$ is their
answer, and we have three possible values for $X$: red ($x_1$), green
($x_2$) and blue ($x_3)$.  The probability that they say red is
$p(X=x_1)$ which is $p_1$, the probability that they say green is
$p(X=x_2)$ which is $p_2$, and the probability that they say blue
$p(X=x_3)$ which is $p_3$.  We can further say that $p_1+p_2+p_3=1$
because every person has to answer with one or another colour.

## Numerical random  variables

Very often, the values that a random variable $X$ may take are numbers
that may be added or subtracted.  For example, we may measure the
height or weight of a sequence of people (or their income, distance
they commute to work, the length of their hair, their age, their blood
pressure, etc etc etc).

We may calculate the expected value $\mathbb{E}(X)$ using the formula

\[
\mathbb{E}(X)=\sum_{i=1}^n x_ip(X=x_i)
\]

The expected value of $X$ (also called the expectation of $X$) is the
long-run average of the amount represented by $X$.  It is easiest to
think of $X$ as an amount of money, as in the following example.
Suppose you play a fruit machine (a gambling device common in pubs in
the UK).  The random variable $X$ is in this case the amount of money
given by the machine as a prize. Random variable $X$ may take three
values: $\$ 0$ with probability 0.9, $\$ 1$ with probability 0.09 and
$\$ 10$ with probability 0.01.  We would write $x_1=\$ 0, x_2=\$ 1,
x_3=\$ 10$ and we would have

\[
\mathbb{E}(X) = x_1*p(X=x_1)+x_2*p(X=x_2)+x_3*p(X=x_3)=\$ 0*0.9 + \$1*0.09 + \$10*0.01 = \$0+\$0.09+\$0.01=\$0.10
\]

Thus you would expect to win $\$0.10$ per play or 10 cents.


