---
output:
  pdf_document: default
  html_document: default
---
# The Binomial distribution

This week we will be discussing the binomial distribution, which has
many and varied applications throughout statistics.


## Bernoulli trials 

A *Bernoulli trial* is the simplest random variable: it amounts to an
experiment with two possible outcomes, conventionally named "success"
and "failure".  Although these are standard words, there is no
associated value judgement and success is not necessarily more
desirable than failure. 

### Examples of Bernoulli trials


The standard example is tossing a coin, with heads recorded as a
success and tails as a failure; but there are loads of examples:
    
*  A patient has a disease and the patient surviving = success,  dies = failure
*  A baby is born; male = success, female = failure
*  Observe the weather on a particular day, rain = failure, not raining=success
*  Throwing a six-sided die and six = success, any other result =  failure
*  Catch a beetle and if it is longer than 2cm this is success,  otherwise failure
*  In a club, ask someone to dance.  s/he says "yes" = success, "no" = failure


Each one of these has a probability of success, usually written $p$.
Of course, the probability of a (fair) coin landing heads has
$p=\frac{1}{2}$ and the probability of a die landing ``6'' has
$p=\frac{1}{6}$.  For some of the other examples above, the
probability is harder to specify but we could take a good guess.

## Simulating Bernoulli trials

We can simulate a series of Bernoulli variables using the
```rbinom()``` function of R.  Remember that you can type
```?rbinom``` at the R prompt to get help on this function.  The
following command shows how to simulate 100 trials with probability
0.5:
    
```{r}
rbinom(100,1,0.5)
```

In the above, the R output is coded in terms of "0" for a failure and
"1" for a success (in the ```rbinom()``` command, the _arguments_ are
the three numbers 100,1,0.5.  The 100 means to simulate 100 trials,
the "1" means to toss the coin once, and the 0.5 gives the probability
of success).  See how about half the trials are successes.


We can do the same thing again:

```{r}
rbinom(100,1,0.5)
```

See how the results are different because R is providing random
numbers.  We can vary the probability of success by changing the third
argument:
    
```{r}
rbinom(100,1,0.1)
```

See how there are much fewer successes because the probability of
success is now only 0.1, or 10\%.  We can make the probability higher:

```{r}
rbinom(100,1,0.9)
```

And now most of the trials are successes. As stated above, the first
argument is the number of trials, and we can play with this by
changing it to, say, 200:

```{r}
rbinom(200,1,0.5)
```

This corresponds to giving a fair coin to each of 200 undergraduates,
telling each of them to toss the coin once, and write down whether the
coin landed heads or tails (but of course there are other
interpretations\ldots).  We might have a weather diary in which we
write down whether it is raining or not for each of 200 days; we might
catch 200 birds and write down whether the bird is male or female, and
so on.

## Multiple Bernoulli Trials: Binomial Distribution

It is possible to be a bit more efficient and tell the 200
undergraduates to toss the coin more than once each.  For example, we
might ask them to toss the coin 5 times.  To do this, we need to
change the second argument from 1 to 5:
    
```{r}
rbinom(100,5,0.5)
```

The R output now shows the number of successes (heads) that each
undergraduate obtained in his or her 5 coin tosses.  Most undergrads
score 2 or 3 heads, but some have as many as 5 and some have zero.  We
would call this ``the binomial distribution with size 5 and
probability 0.5'' as there are five trials (coin tosses) and the
probability of each one being a success is 0.5.  This can also be
written $\mathrm{Bin}(5,0.5)$.

## Changing the parameters
    

It is possible to change the probability of success.  Suppose we give
the undergraduates a biased coin with a 90\% probability of landing
heads.  To simulate this, we need to change the third argument to 0.9,
or 90\%:
    
```{r}
rbinom(100,5,0.9)
```

Now most of the students have 5 out of 5 successes, which is not
surprising because of the high probability of success.


## Tabulating results of random simulations

It is possible to make a table of the results using the ```table()```
function in R.  To simulate 100 students each tossing a coin 5 times
with probability of heads = 0.5, we can use the following:

```{r}
set.seed(0)
x <- (rbinom(100,5,0.5))
table(x)
```

(the ```set.seed()``` command ensures reproducible result).  Thus, out
of 100 students, 2 scored 0 heads out of 5 tosses, 11 scored 1 head,
39 scored 2 heads, and so on.

## Increasing the number of trials

We can try the same thing with a larger number of students:

```{r}
x <- rbinom(1000,5,0.5)
table(x)
```

Now 27 students out of 1000 have zero heads out of 5 tosses, 167 have
one head, and so on.


## Plotting the results

It is possible to plot this table in R; see figure
  \@ref(fig:binomialplot).  Remember that you may type ```?plot``` at
  the R prompt to get help.

```{r binomialplot,fig.cap="Tabulated observations from the binomial distribution", echo=FALSE,message=FALSE,warning=FALSE}
plot(table(x))
```

## Mathematics of the binomial distribution


Let $X$ be a binomial random variable with $n$ trials and probability
of success $p$.  There is a formula for the probability of scoring
each number of heads.

It is:


\[
P(X = r) = \frac{n!}{r!(n-r)!}p^r(1-p)^{n-r} \qquad r = 0, 1, \ldots, n
\]

where

* $n$ is the number of trials
* $r$ the number of successes we want the probability of occurring
* $p$ is the probability that an individual trial is a success


Here, "$n!$" is pronounced "$n$ factorial" and is the number of
ways of arranging $n$ objects.  It is given by $n!=1\times 2\times
3\times\cdots\times n$; you multiply all the whole numbers up to $n$
together.  Thus $6!=1\times 2\times 3\times 4\times 5\times 6=720$.
You may see this formula written slightly differently in the
literature:

\[
{n\choose r} p^r\left(1-p\right)^{n-r}
\]

where $n\choose r$ is pronounced ``n choose r'' and
means $\frac{n!}{r!(n-r)!}$.


Let $X$ be a binomial random variable representing the number of
successes from $n$ trials.  Suppose we want to calculate the
probability that a student obtains $r=2$ successes from $n=5$ trials,
each with probability $p=0.3$ of success.  Then we get:
    
\[P(X = 2) = \frac{5!}{2!(5-2)!}0.3^2(1-0.3)^{5-2}\]
    
```{r}
factorial(5)/(factorial(2)*factorial(3)) *  0.3^2 * (1-0.3)^(5-2)
```

which means a probability of a little over 30\%.  We can calculate
this more simply using the ```dbinom()``` function:
    
    
```{r}
dbinom(2,5,0.3)
```
    
which gives the same answer.

## Cumulative Distribution Function

Sometimes we will want to compute the probability that a binomial
random variable is less than or equal to some value. To do this we can
use the (cumulative) distribution function or cdf.

The formula is:

\[
P(X \leq r) = 
\sum_{i=0}^r
\frac{n!}{i!(n-i)!}p^i(1-p)^{n-i} \qquad r = 0, 1, \ldots, n
\]


Suppose we want to calculate the probability that a student obtains
$r=2$ or fewer successes from $n=5$ trials, each with
probability $p=0.3$ of success.

    
\[P(X \leq 2) = \sum_{i=0}^2\frac{5!}{i!(5-i)!}0.3^i(1-0.3)^{5-i}\]

Then we get:
    
```{r}
factorial(5)/(factorial(0)*factorial(5)) * 0.3^0 * (1-0.3)^(5-0) +
factorial(5)/(factorial(1)*factorial(4)) * 0.3^1 * (1-0.3)^(5-1) +
factorial(5)/(factorial(2)*factorial(3)) * 0.3^2 * (1-0.3)^(5-2)
```

This is equivalent to 

```{r}
dbinom(0,5,0.3) + dbinom(1,5,0.3) + dbinom(2,5,0.3)
```

These three probabilities can be computed in one line:

```{r}
dbinom(0:2,5,0.3)
```

We can add these three probabilities to obtain the same answer as above

```{r}
sum(dbinom(0:2,5,0.3))
```


Or, more simply:
    
    
```{r}
pbinom(2,5,0.3)
```
    
which gives the same answer.

## The binomial distribution in R


In general, a random distribution (here the binomial) has four R
functions associated with it:

* ```dbinom()``` Probability mass function, $P(X = x)$
* ```pbinom()``` Cumulative distribution function, $P(X \leq x)$
* ```rbinom()``` Random sample from the binomial distribution
* ```qbinom()``` Quantile function

(sometimes one or more of these functions may be absent, typically if
the distribution is very complicated or difficult to work with; but
mostly all four will be present and documented together).


### Example 

Consider Bin(10,0.3), that is the binomial distribution with $n=10$
and $p=0.3$.

Then we can use the four functions in turn:

```{r}
dbinom(5,10,0.3)   # probability that we observe r=5 successes
pbinom(4,10,0.3)   # probability that we observe r<=4 successes
rbinom(20,10,0.3)  # 20 random observations from Bin(10,0.3)
qbinom(0.99,10,0.3) # 99th percentile of Bin(10,0.3)
```

## Numerical verification

It is always a good idea to try and verify theoretical results using
random sampling.  We will take one million observations (that is, one
million undergraduates, each tossing a coin 5 times) and recording the
number of heads for each undergraduate:

```{r}
table(rbinom(1000000,5,0.3))
```


So, out of a million observations we have exactly 309666 students with
two heads, a proportion of 0.309666, not too far from the exact value
of 0.3087 (the slight difference is caused by random variation).

We can see all the probabilities in one line using the colon construction:
  
```{r}
dbinom(0:5,5,0.3)
```

and we should check that these probabilities add to one:
  
```{r}
sum(dbinom(0:5,5,0.3))
```

Using this, we can check any binomial distribution:

```{r}
sum(dbinom(0:115,115,0.324324))
```

as expected.

## Expected Value and Variance

You will recall the discussion about variance and standard deviation
from  week 2.  The Binomial distribution has probability mass function

\[\frac{n!}{r!(n-r)!}p^r(1-p)^{n-r}.\]



* The *mean value* of the binomial distribution  $E[X] = np$
* The *variance* of the binomial distribution  $\text{Var}[X] = np(1-p)$
* The *standard deviation* of the binomial distribution $\text{Std}[X] =\sqrt{np(1-p)}$.

We will now verify these formulas using direct numerical sampling.

### Example

Consider $n=7$ trials with probability $p=0.3$ of success.  The
formulas give a mean of $7\times 0.3=2.1$ and a standard deviation of
$\sqrt{7\times 0.3\times 0.7}\simeq 1.21$.  Compare these with the
values obtained from numerical sampling:

```{r}
mean(rbinom(1e6,7,0.3))
var(rbinom(1e6,7,0.3))
```

The numerical answers are close to the theoretical values in both
cases.

## More trials and further R functions

So far we have used 5 trials per observation (we say that the *size*
of the distribution is 5), but there is no reason why we cannot use
more.  We can try, for example, 10 trials:


```{r}
rbinom(100,10,0.4)
```

this would be Bin(10,0.4).  We could try higher numbers, say 30:

    
```{r}
rbinom(100,30,0.4)
```

for a Bin(30,0.4) distribution.  We can plot the density of the
binomial distribution with size 30 and probability 0.4 using the
```dbinom()``` function directly:

```{r binplot,fig.cap="The binomial distribution", echo=FALSE,message=FALSE,warning=FALSE}
plot(0:30,dbinom(0:30,30,0.4),type='h')
```

Suppose we ask a slightly different question.  We might be looking at
a Bin(30,0.4) distribution and wonder what the probability of
obtaining 15 or more successes, i.e. $P(X \geq 15)$.  We can calculate
the probability of obtaining 15,16,17,\ldots successes using the
```dbinom()``` command:


```{r}
dbinom(15:30,30,0.4)
```

Rather than adding these up with a calculator we can use R to do the addition:
    
    
```{r}
sum(dbinom(15:30,30,0.4))
```

Giving us an answer of about 17.5\%.  We can also use the cdf command
directly. Notice that $P(X \geq 15) = 1 - P(X \leq 14)$.

```{r}
1-pbinom(14,30,0.4)
```


We can verify this numerically:
    
```{r}
sum(rbinom(1e6,30,0.4) >= 15)/1e6
```

showing close agreement between numerical and theoretical values.

## Example



Question: Suppose a scientist catches 10 fish of a particular species,
and it is known that these fish have a 30\% probability of being
female at this time of year.

What type of random variable can be used to model this
scenario?

Answer: A binomial random variable with size 10 and probability of
``success'' 0.3.  We can write $X \sim \operatorname{Bin}(10, 0.3)$.

Question:  What is the probability that exactly 5 are female?

Solution:

```{r}
dbinom(5,10,0.3)
```

Question: What is the probability that 6 or fewer fish are female?
 
```{r}
pbinom(6,10,0.3)
```
What is the probability that at least one fish of each gender is caught?

We need $P(1 \leq X \leq 9) = 1 - P(X = 0) - P(X = 10) = 0.972$.


```{r}
sum(dbinom(1:9,10,0.3))
1-dbinom(0, 10, 0.3)-dbinom(10, 10, 0.3)
```


What is the expected value and standard deviation for number of female fish caught?

$E[X] = np = 10 \times 0.3 = 3$

$\text{Std}[X] = \sqrt{np(1-p)} = \sqrt{10 \times 0.3 \times 0.7} = 1.449138$

```{r}
10*0.3
10*0.3*(1-0.3)
```

Numerically verify the answers to the questions above.

```{r}
#P(X = 5)
sum(rbinom(1e6,10,0.3)==5)/1e6

#P(X <=6)
sum(rbinom(1e6,10,0.3)<=6)/1e6

#P(1 <= X <=9)
x <- rbinom(1e6,10,0.3)
sum((x>0) & (x<10))/1e6

#E[X] and Std[X]
mean(rbinom(1e6,10,0.3))
sd(rbinom(1e6,10,0.3))
```
