# Confidence intervals

Rather than estimating a parameter with a single value, sometimes it
is better to provide a range of values, which we are "fairly certain"
will contain the parameter of interest.  When these intervals are
constructed in a particular way, they are called *confidence
intervals*.

We want to give an upper and lower bound for the population mean; we
specify that we want to be "95\% sure" that the true value is between
our two bounds.  We have seen confidence intervals used in this course
previously but only in an informal way.  In this chapter we will give
the precise definition of confidence interval.

## Confidence Intervals for the Population Mean with a Known Population Variance

Suppose we have observations $x_1,\ldots, x_n$ drawn from a Normal
distribution $N(\mu,\sigma)$, where the standard deviation $\sigma$ is
known.  To obtain a 95\% confidence interval for a mean, with a known
population variance, we can compute these bounds as follows.  We know
that the sample mean is normally distributed with mean $\mu$ and
standard deviation given by the standard error of the mean
$\sigma/\sqrt{n}$, i.e. $\bar{X} \sim N(\mu, \sigma/\sqrt{n})$.

If we construct an interval from $\overline{x} +
\alpha\frac{\sigma}{\sqrt{n}}$ to $\overline{x} -
\alpha\frac{\sigma}{\sqrt{n}}$ then this will (in very loose language)
"capture" the true value of the mean $\mu$ with a certain probability.
We want this probability to be 0.95, or 95\%, so we choose $\alpha$ to
be `qnorm(0.975)` (that is, about 1.96) which leaves 2.5\%
probability on either side, as in figure \@ref(fig:confidenceconcept).


```{r confidenceconcept,fig.cap="Conceptual diagram of confidence interval", echo=FALSE}
x <- seq(from=-4,to=4,len=100)
plot(x,dnorm(x),type="l")
xp <- qnorm(0.975)
xx <- seq(from=xp,to=4,len=100)
polygon(c(xx,rev(xx)),c(dnorm(xx),xx*0),border=NA,col='red')
xx <- -xx
polygon(c(xx,rev(xx)),c(dnorm(xx),xx*0),border=NA,col='red')
text(0,0.1,"95%")
text(+xp+0.7,0.1,"2.5%")
text(-xp-0.7,0.1,"2.5%")
```


## Example: weight of bread.

Suppose a government inspector buys 30 loaves of bread from a bakery
and wants to know the mean weight of all loaves sold by that bakery.  The data is, in grams:

```{r}
loaves <- c(804, 804, 800, 797, 805, 800, 810, 803, 806, 800, 801, 802, 
800, 808, 806, 804, 810, 799, 804, 803, 805, 803, 804, 793, 808, 
810, 801, 802, 797, 802)
```

*Question*: Give a 95\% confidence interval for the population mean.

*answer* The sample mean and standard deviation is

```{r}
mean(loaves)
sd(loaves)
```

and the standard error of the mean is

```{r}
SEM <- sd(loaves)/sqrt(30)
SEM
```

For a 95\% confidence interval we find the value of that has 2.5\% (=
half of 5\%) probability to the right, that is 97.5\% to the left:

```{r}
qnorm(0.975)
```

and the 95\% confidence interval would be given by

```{r}
c(mean(loaves) - qnorm(0.975)*SEM, mean(loaves) + qnorm(0.975)*SEM)
```

To get a 99\% confidence interval instead of the default 95\%, we have
exactly the same construction except we need to use
`qnorm(0.995)`:

```{r}
c(mean(loaves) - qnorm(0.995)*SEM, mean(loaves) + qnorm(0.995)*SEM)
```

See how this interval is wider than the 95\% interval: the price one
pays for more certainty is lower precision.  

## Confidence intervals with the Student $t$-distribution

The previous example had enough observations to be able to estimate
the population standard deviation reasonably.  If that is not the
case, then we have to use methods deriving from the Student $t$-test.
R can do this easily.

### Example: heart rates.

A medical scientist takes six patients and measures their heart rate.  The data is:

```{r}
heart <- c(66,73,72,80,84,60)
```

Calculate a 95\% confidece interval for this data.

There are only six datapoints, so the $Z$-score method is
inapplicable.  The `t.test()` in R gives:

```{r}
t.test(heart)
```

So the 95\% confidence interval is about $(63.3, 81.7)$.  It is quite
wide because we have so few observations.  Note that the Z-score
method can also be used:

```{r}
c(
mean(heart) - qnorm(0.975)*sd(heart)/sqrt(6),
mean(heart) + qnorm(0.975)*sd(heart)/sqrt(6)
)
```

This gives a confidence interval that is much narrower than given by
the $t$-test, and should not be used.

## Frequentist interpretation of the confidence interval

The value of a confidence interval is that it contains the true value
of the mean with a high probability.  Figure \@ref(fig:confidencefreq)
shows repeated realisations of a confidence interval where the true
mean of zero is indicated as a vertical line.  Most of the confidence
intervals include the true value of zero; those that do not are shown
in red.  The inference is that a particular confidence interval
probably contains the true parameter, although the correct definition
(the set of points $x$ for which $H_0\colon\mu=x$ would not be
rejected) is problematic.

```{r confidencefreq,fig.cap="Repeated confidence intervals with intervals missing the true value shown in red", echo=FALSE}
set.seed(1)
par(lend=1)
f <- function(x){t.test(x)$conf.int}
plot(NA, xlim=c(-2,2),ylim=c(0,70),xlab="",ylab="",axes=F)
for(i in 1:70){
  jj <- f(rnorm(5))
  if(prod(jj)<0){col <- "black"} else {col <- "red"}
  segments(x0=jj[1],x1=jj[2],y0=i,lwd=2,col=col)
}
abline(v=0)
```
