# Confidence intervals

Rather than estimating a parameter with a single value, sometimes it
is better to provide a range of values, which we are "fairly certain"
will contain the parameter of interest.  When these intervals are
constructed in a particular way, they are called *confidence
intervals*.

We want to give an upper and lower bound for the population mean; we
specify that we want to be "95\% sure" that the true value is between
our two bounds.  We have seen confidence intervals used in this course
previously but only in an informal way.

It is easiest to approach confidence intervals through a specific
example: finding the population mean of Gaussian observations with
unknown standard deviation; we will use a significance level of $0.05$
[this is usually expressed as "95% confidence", for we are interested
in the remainder $1-0.05=0.95$].  Suppose we have observations `1.5,
2.3, -0.7, 1.4, 1.2`.  We might test, in turn,
$\mu=-3,\mu=-2,\ldots,\mu=2,\mu=3$:

```{r}
options(digits=3)
x <- c(1.5, 2.3, -0.7, 1.4, 1.2)
c(
`-3` = t.test(x,mu=-3)$p.value,
`-2` = t.test(x,mu=-2)$p.value,
`-1` = t.test(x,mu=-1)$p.value,
 `0` = t.test(x,mu= 0)$p.value,
`+1` = t.test(x,mu=+1)$p.value,
`+2` = t.test(x,mu=+2)$p.value,
`+3` = t.test(x,mu=+3)$p.value)
```

We see from the above R code that we reject $\mu=-3,-2,-1$; fail to
reject $\mu=0,+1,+2$, and reject $\mu=+3$.  We can infer that
$\mu=0,+1,+2$ are "reasonable" values for $\mu$, as we fail to reject
nulls for these values of $\mu$.  In general, suppose we test a null
$H_0\colon\mu=\mu_0$, and fail to reject $H_0$, then we say that
$\mu_0$ is inside the confidence interval.  Confidence intervals are
given by `t.test()`:


```{r}
t.test(x)
```

Above we see that the confidence interval is `(-0.239,2.519)`.  Thus
we would reject any hypothesis that the mean is less than `-0.239` or
greater than `2.519`.


## Example: heart rates.

A medical scientist takes six patients and measures their heart rate.
The data is:

```{r}
heart <- c(66,73,72,80,84,60)
```

Calculate a 95\% confidence interval for this data.  The `t.test()` in
R gives:

```{r}
t.test(heart)
```

So the 95\% confidence interval is about $(63.3, 81.7)$.  It is quite
wide because we have so few observations.  We would reject any
hypothesis that the true mean heart rate is outside this range.  For
example, we would reject $H_0\colon\mu=63$ as this is less than the
lower limit of 63.3.

## Confidence intervals for normal data with known variance

If we know the variance of our observations, this makes our life a
little easier.  Suppose we have observations $x_1,\ldots, x_n$ drawn
from a Normal distribution $N(\mu,\sigma)$, where the standard
deviation $\sigma$ is known.  To obtain a 95\% confidence interval for
a mean, with a known population variance, we can compute these bounds
as follows.  We observe that
$Z=\frac{\bar{x}-\mu}{\sigma/\sqrt{n}}\sim N(0,1)$.  If we choose a
5\% level of significance, then because `pnorm(0.975) = 1.96` (figure
\@ref(fig:confidenceconcept)) we fail to reject $\left|Z\right|\leq
1.96$, which is the basis of our confidence interval: we would fail to
reject any $\mu$ in the range

\[
\overline{x}\pm 1.96\frac{\sigma}{\sqrt{n}}
\]


(just invert the formula for $Z$ using algebra).

```{r confidenceconcept,fig.cap="Conceptual diagram of confidence interval", echo=FALSE}
x <- seq(from=-4,to=4,len=100)
plot(x,dnorm(x),type="l")
xp <- qnorm(0.975)
xx <- seq(from=xp,to=4,len=100)
polygon(c(xx,rev(xx)),c(dnorm(xx),xx*0),border=NA,col='red')
xx <- -xx
polygon(c(xx,rev(xx)),c(dnorm(xx),xx*0),border=NA,col='red')
text(0,0.1,"95%")
text(+xp+0.7,0.1,"2.5%")
text(-xp-0.7,0.1,"2.5%")
```


## Example: weight of bread.

Suppose a government inspector buys 30 loaves of bread from a bakery
and wants to know the mean weight of all loaves sold by that bakery.  The data is, in grams:

```{r}
loaves <- c(804, 804, 800, 797, 805, 800, 810, 803, 806, 800, 801, 802, 
800, 808, 806, 804, 810, 799, 804, 803, 805, 803, 804, 793, 808, 
810, 801, 802, 797, 802)
```

*Question*: Give a 95\% confidence interval for the population mean.

*answer* The sample mean and standard deviation is

```{r}
mean(loaves)
sd(loaves)
```

and the standard error of the mean is

```{r}
SEM <- sd(loaves)/sqrt(30)
SEM
```

For a 95\% confidence interval we find the value of that has 2.5\% (=
half of 5\%) probability to the right, that is 97.5\% to the left:

```{r}
qnorm(0.975)
```

and the 95\% confidence interval would be given by

```{r}
c(mean(loaves) - qnorm(0.975)*SEM, mean(loaves) + qnorm(0.975)*SEM)
```

To get a 99\% confidence interval instead of the default 95\%, we have
exactly the same construction except we need to use
`qnorm(0.995)`:

```{r}
c(mean(loaves) - qnorm(0.995)*SEM, mean(loaves) + qnorm(0.995)*SEM)
```

See how this interval is wider than the 95\% interval: the price one
pays for more certainty is lower precision.  


## Frequentist interpretation of the confidence interval

The value of a confidence interval is that it contains the true value
of the mean with a high probability.  Figure \@ref(fig:confidencefreq)
shows repeated realisations of a confidence interval where the true
mean of zero is indicated as a vertical line.  Most of the confidence
intervals include the true value of zero; those that do not are shown
in red.  The inference is that a particular confidence interval
probably contains the true parameter, although the correct definition
(viz, the set of points $x$ for which $H_0\colon\mu=x$ would not be
rejected) is problematic.

```{r confidencefreq,fig.cap="Repeated confidence intervals with intervals missing the true value shown in red", echo=FALSE}
set.seed(1)
par(lend=1)
f <- function(x){t.test(x)$conf.int}
plot(NA, xlim=c(-2,2),ylim=c(0,70),xlab="",ylab="",axes=F)
for(i in 1:70){
  jj <- f(rnorm(5))
  if(prod(jj)<0){col <- "black"} else {col <- "red"}
  segments(x0=jj[1],x1=jj[2],y0=i,lwd=2,col=col)
}
abline(v=0)
```
