# ANOVA

Consider the following dataset:

```{r}
light <-
data.frame(experiment=
letters[c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,2,2,2,2,2,2,2,2,2,
2,2,2,2,2,2,2,2,2,2,2,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,3,4,
4,4,4,4,4,4,4,4,4,4,4,4,4,4,4,4,4,4,4,5,5,5,5,5,5,5,5,5,5,5,5,5,
5,5,5,5,5,5,5)],speed=c(85,74,90,107,93,85,95,98,98,88,100,98,93,
65,76,81,100,100,96,96,96,94,96,94,88,80,85,88,90,84,83,79,81,88,
88,83,80,79,76,80,88,88,88,86,72,72,62,86,97,95,88,91,85,87,84,84,
85,84,84,84,89,81,81,82,80,77,76,74,75,76,91,92,89,86,88,72,84,85
,85,78,89,84,78,81,76,81,79,81,82,85,87,87,81,74,81,94,95,80,81,87))

head(light)
```

This is a modified version of a dataset documented at `?morley` in
the R system.  It shows the speed of light as measured in five
different experiments in the nineteenth century.  The speed column
shows the discrepancy between the measured speed of light and a
standard benchmark speed.  Is there any evidence that the experiments
have systematic bias?

To answer this question we first plot the data:

```{r}
boxplot(speed~experiment,data=light,col="gray")
```

In the figure above, there is some suggestion of a difference between
the five experiments; but is this difference real, or can it be
explained by random variability?  To answer this, we set up a null
hypothesis:

\[
H_0\colon\mu_A=\mu_B=\mu_C=\mu_D=\mu_E
\]

(that is, the mean speed of light is the same for all five
experiments).  We need to calculate a $p$-value for this null, and
this is provided by the `aov()` function ("AOV" stands for
"Analysis of variance"):
 
```{r} 
summary(aov(speed~experiment,data=light))
```

Note that we take the summary of the result of `aov()` as this
gives us more information.  Although the output is a little cryptic,
we can see the $p$-value for our null is 0.0031, well below the 5\%
threshold.  We may reject the null hypothesis and conclude that the
experiments exhibited bias.

It is possible to use the `lm()` function, which gives us slightly
different information:

```{r}
summary(lm(speed~experiment,data=light))
```

The format is very similar to that of scatterplot linear regression
but the "intercept" is experiment "a" (R puts the factors in
alphabetic order) and the other lines give the difference between
experiment= "a" and the other experiments "b" through "e".  The
overall $p$-value of 0.00311 is the same as in `aov()`, because it
refers to the same null.

