# Fisher's exact test {#fisher}

Suppose I am interested in the prevalence of left handedness and wish
to determine whether there is a difference between boys and girls.  I
might survey some students and end up with a dataset like the
following:


```{r}
a <- matrix(c(5,2,6,14),2,2,byrow=TRUE)   # define matrix 
dimnames(a) <- list(gender=c("M","F"),lefthanded=c(T,F))  # display only
a
```

Such a dataset is known as a *contingency table* and is surprisingly
common in statistics.  I would say that over half of the statistical
analyses I perform for clients is on contingency tables like the one
above.

Is there any evidence that handedness differs by gender?  Just looking
at the table, we can see that the majority of males are left handed
and the majority of females are right handed.  We need to provide
statistical analysis of this question which means providing a p-value.

The starting point would be to provide a null hypothesis, which is
that handedness is independent of gender.  We would thus have
$H_0\colon p(L|M)=p(L)$ (in words, "the probability of being left
handed, given that you are male, is equal to the probability of being
left handed").  This is mathematically identical to the assertion that
$p(L|M)=p(L|F)$.  It certainly looks as though there is a difference,
but to quantify this we need a more careful analysis.

## The hypergeometric distribution

Suppose we have an urn with $m$ white and $n$ black balls, and we draw
out $k$ balls and write down their colour.  Note carefully that we are
drawing *without replacement*: we draw a ball from the urn and then do
not put it back in the urn (if we were to replace the ball in the urn,
the number of white balls would be binomial with size $k$ and
probability $\frac{m}{m+n}$).

What is the distribution of the number of white balls drawn without
replacement?  Well, it has a particular distribution known as the
*hypergeometric distribution*.  This is similar to the binomial
distribution in that it is discrete, non-negative, and has a known
maximum.  The R idiom for this distribution is ```dhyper()``` which
calculates the density function; we will not need this because R
calculates the $p$-value for us.  We can make the following mapping
between drawing balls out of a urn and left/right male/female dataset
as follows:

* white ball $\longrightarrow$ left handed
* black ball $\longrightarrow$ right handed
* drawn from urn $\longrightarrow$ male
* left in urn $\longrightarrow$ female

Alternatively we might say 

* white ball $\longrightarrow$ male
* black ball $\longrightarrow$ female
* drawn from urn $\longrightarrow$ left handed
* left in urn $\longrightarrow$ right handed

Taking the second  of the two identifications, we have

* number of white balls = number of males =5+2=7
* number of black balls = number of females = 6+14=20
* drawn from urn = left handed = 5+6=11
* left in urn = right handed = 2+14=15

With this identification, the *observation* is the number of white
balls drawn from the urn (= the number of left handed males) which is
5.

Recall the definition of p-value: "the probability, if the null is
true, of obtaining the observation or an observation more extreme".
In this case (if we use a one-sided test), "more extreme" means
greater than the observation of 5; and we know that the number of left
handed males (=white balls drawn) cannot be larger than 7.  We can get
a visual impression easily, using R:

```{r}
x <- 0:7  # define x-axis
plot(x,dhyper(x,7,20,11),type='h',lwd=4,col=c(rep("black",5),rep("red",3))) 
legend("topright",lwd=4, col=c("black","red"),legend=c("regular points","more extreme"))
```

Fortunately, R can deal with all this with the ```fisher.test()``` function:

```{r}
fisher.test(a, alternative="greater")
```

(recall that R variable ```a``` was defined at the beginning of the
chapter).  We can see from the $p$-value of about 7\% that the result
is not significant: there is no evidence for a difference in
handedness between male and female students.

## Other applications

Fisher's exact test can be applied whenever we have two different ways
of classifying subjects.  We used left/right handed and male/female
above, but the same thinking applies to a very wide range of datasets.
I include some examples below.  


### Medical research: old treatment, new treatment.

*Question* Suppose we are researching a new medical treatment and wish
to discover whether it is better than the old one.  We take some
patients and treat 6 with the old treatment and 8 with the new; of the
6 patients treated with the old treatment, 2 are cured.  And of the 8
treated with the new treatment, 5 were cured.  Is this evidence for the
new treatment being better than the old?

*answer*:

The first step is to translate the information given into R idiom.  In
this case we have a two-row matrix with the top row corresponding to
the old treatment and the bottom row corresponding to the new
treatment.  The first column corresponds to cured patients and the
second column corresponds to non-cured patients.

```{r}
M <- matrix(c(2,5,4,2),2,2)
M
```

The above is fine but is easier to understand if we modify ```M``` so
that the rows and columns are named.  The R idiom to do this is nice,
but not part of the learning objectives of the course and you will not
be tested on it.  It is just cosmetic.


```{r}
dimnames(M) <- list(treatment=c("new","old"),cured=c(TRUE,FALSE))
M
```

In the above, the matrix ```M``` is unchanged but has new information
in the rownames and columnnames.  

The second step is to formulate a null: $H_0\colon
p(\mathrm{cure}|\mathrm{new}\,\mathrm{treatment}) = p(\mathrm{cure})$.
We need a $p$-value for this null:


```{r}
fisher.test(M,alternative="less")
```

(note the use of a one-sided test: we are testing for the new
treatment being *better* than the old one, not *different* from it).
The $p$-value exceeds 5\%, so this indicates that the improvement
could easily have occurred by chance.  There is no evidence that the
new treatment is any better than the old.