---
title: "matpmda_exam_11_may_2026.Rmd"
author: "NAME AND STUDENT ID GO HERE"
date: "2025-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This exam has FIVE questions.  Answer them all.  Each will be marked
according to the common marking scheme and your overall mark will be
the arithmetic mean of the five marks.


# Question 1

A sports scientist collects strength data from two groups of students.
The first group is a group of athletes who might be expected to be
strong, and the second group is a control group of nonathletic
students.  The first group has strength ratings of:

```
c(135.6, 140.7, 133, 140.4, 127.4, 133.3, 128.8, 144.1)
```

And the second group has strength ratings of:

```
c(115.1, 125.4, 122, 116.2, 123.3, 126.6)
```

Is there evidence that the first group is actually stronger than the second?

* State a sensible null hypothesis
* State the precise definition of p-value and explain
     what “more extreme” means in this context
* Is a one-sided or two-sided test needed? justify
* Perform a student t-test using R and interpret



***Answer here***

Null: both groups have identical mean stengths.

P-value: "the probability, if the null is true, of obtaining the
observation or an observation more extreme".  Here, the observation is
the difference in mean strengths and "more extreme" means a larger
difference than observed.

One-sided, athletes known to be stronger

```{r}
x <- c(135.6, 140.7, 133, 140.4, 127.4, 133.3, 128.8, 144.1)
y <- c(115.1, 125.4, 122, 116.2, 123.3, 126.6)
t.test(x,y,alternative="greater")
```

small pvalue indicates strong evidence that the athletes are indeed
stronger.

# Question 2

Random variable X is drawn from a normal distribution with mean 1.1
and standard deviation 2.

* Calculate the probability of X being less than -0.5
* What is the probability of X lying between -0.8 and 0.46?
* Verify your answers to parts (a) and (b) above using numerical sampling.
* (Harder) Random variable Y is independent of X and drawn from the
  same distribution.  Estimate the probability that $X^2 + XY$ is less
  than or equal to 1.5.

***Answer here***


```{r}
pnorm(-0.5, 1.1, 2)
diff(pnorm(0.46, -0.8), 1.1, 2)
x <- rnorm(1e6, 1.1, 2)
table(x < -0.5)/1e6
table( (x > -0.8) & (x < 0.46))/1e6
y <- rnorm(1e6, 1.1, 2)
table(x^2 + x*y <= 1.5)
```

# Question 3

It is known that seeds from a particular type of plant germinate with
a probability 0.2.  An environmental scientist visits a location,
suspected of having seeds that are less likely to germinate than
averate.  She samples 325 seeds from this species and finds that 55
germinate.

* State a sensible null hypothesis.
* Under the null, what is the expected number of germinations?
* State the precise definition of pvalue and say what "more extreme" means in this context.
* Calculate a p-value for the observation of 55 germinations.  Interpret your finding.
* (harder) Using these figures (and the formula $Z = (x-\mu)/\sigma$,
  what is the Z-score for the observation?  Use pnorm() to calculate a
  pvalue under the normal approximation.

***Answer here***

Null: this location has the same probability of a seed germinating as
anywhere else.

```{r}
0.2*325
```

P-value: "the probability, if the null is true, of obtaining the
observation or an observation more extreme".  Here, the observation is
55 and more extreme means less than this.

```{r}
sum(dbinom(0:55,325,0.2))
```

pvalue exceeds 0.05, no evidence.

```{r}
pnorm(55, 325*0.2, sqrt(325*0.2*0.8))
```



# Question 4

A Students are expected to have higher grades if they spend more time
studying.  An educational theorist collects data on 17 students and the
number of hours they spend studying is:

```{r}
hours <-
c(20.3, 18.6, 19.5, 20.3, 18.7, 21.1, 19, 22.4, 18.6, 18.2, 20.9,
18.8, 19.1, 22.7, 19.1, 20.2, 19.6)
hours
```

Their grade average is:

```{r}
grades <-
c(7.7, 6.7, 9.6, 10, 9, 10.9, 8.8, 9.2, 7.8, 6.1, 10.4, 9.9,
8.4, 9.5, 8.3, 7.6, 10.2)
grades
```

Is there evidence that students who spend more hours studying score higher grades?

* State a sensible null hypothesis
* State the precise definition of p-value and explain what “more extreme” means in this context
* Is a one-sided or two-sided test needed? justify
* Perform a linear regression using R and interpret

***Answer here***

Null: no relation between study hours and grades.

P-value: "the probability, if the null is true, of obtaining the
observation or an observation more extreme".  Here, the observation is
the fitted slope and more extreme means a larger value.


```{r}
hours <-
c(20.3, 18.6, 19.5, 20.3, 18.7, 21.1, 19, 22.4, 18.6, 18.2, 20.9,
18.8, 19.1, 22.7, 19.1, 20.2, 19.6)
grades <-
c(7.7, 6.7, 9.6, 10, 9, 10.9, 8.8, 9.2, 7.8, 6.1, 10.4, 9.9,
8.4, 9.5, 8.3, 7.6, 10.2)
plot(grades~hours)
fit <- lm(grades~hours)
summary(fit)
abline(fit)
```


# Question 5

A scientist is investigating whether two species of snails have
different responses to moisture.  He takes 5 snails of type A and 7 of
type B and allows them to move freely in a two-chambered enclosure,
with one half being dry and the other being wet.  He finds that 4
snails, of which three are of type A, prefer the dry part of the
container.

* State a sensible null hypothesis
* State the precise definition of p-value and explain what “more extreme” means in this context
* Is a one-sided or two-sided test needed? justify
* Using `fisher.test()`, or otherwise, assess your null and interpret.



***Answer here***

Null: both types of snail have equal responses to wet/dry conditions.

P-value: "the probability, if the null is true, of obtaining the
observation or an observation more extreme".  Here the observation is
one snail of type B in the dry part of the enclosure and more extreme
would be a smaller probability under the null.

two-sided, no direction known.

```{r}
M <- matrix(c(2,3,6,1),2,2)
dimnames(M) <- list(condition=c("wet","dry"), species=c("A","B"))
M
fisher.test(M)
```

Not significant.

