---
title: "matpmda_exam_14_may_2025.Rmd"
author: "NAME AND STUDENT ID GO HERE"
date: "2025-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Question 1

A sports scientist collects strength data from two groups of students. The first group is a
group of athletes who might be expected to be strong, and the second group is a control
group of nonathletic students.
The first group has strength ratings of:

```
c(135.6, 140.7, 133, 140.4, 127.4, 133.3, 128.8, 144.1)
```

And the second group has strength ratings of:

```
c(115.1, 125.4, 122, 116.2, 123.3, 126.6)
```

Is there evidence that the first group is actually stronger than the second?

* State a sensible null hypothesis
* State the precise definition of p-value and explain what “more extreme” means in this
context
* Is a one-sided or two-sided test needed? justify
* Perform a student t-test using R and interpret


***Answer here***


# Question 2

Random variable X is drawn from a normal distribution with mean -1.23 and standard deviation 2.05.

* Calculate the probability of X being less than -3.65.
* What is the probability of X exceeding -2.01?
* What is the probability of X lying between -0.8 and 0.46?
* Verify your answers to parts (a) and (b) above using numerical sampling.
* (Harder) verify your answers to part (c) above using numerical sampling.


***Answer here***


# Question 3

Concrete specimens are assessed using a certain standardized test involving stressing the
specimen in a rig. The concrete either breaks (fails), or does not break (passes the test;
succeeds). It is known that the probability of passing the test is 0.202.
A material scientist claims that painting the concrete with a particular paint increases the
success rate. She takes 325 samples, paints them, and finds that 73 pass.
(a) 

* State a sensible null hypothesis.
*  Under the null, what is the expected number of successes?
*  State the precise definition of pvalue and say what "more extreme" means in this context.
* Calculate a p-value for the observation of 73 successes. Interpret your finding.
* Using these figures (and the formula $Z = (x-\mu)/\sigma$, what is the Z-score for the observation?  Use pnorm() to calculate a pvalue under the normal approximation.

***Answer here***


# Question 4

Students are expected to have higher grades if they spend more time studying. An educational theorist collects data on 17 students and the number of hours they spend studying
is:

```{r}
hours <-
c(20.3, 18.6, 19.5, 20.3, 18.7, 21.1, 19, 22.4, 18.6, 18.2, 20.9,
18.8, 19.1, 22.7, 19.1, 20.2, 19.6)
hours
```


Their grade average is:

```{r}
grades <-
c(7.7, 6.7, 9.6, 10, 9, 10.9, 8.8, 9.2, 7.8, 6.1, 10.4, 9.9,
8.4, 9.5, 8.3, 7.6, 10.2)
grades
```

Is there evidence that students who spend more hours studying score higher grades?

* State a sensible null hypothesis
* State the precise definition of p-value and explain what “more extreme” means in this context
* Is a one-sided or two-sided test needed? justify
* Perform a linear regression using R and interpret

***Answer here***

