---
output: pdf_document
title: The Gaussian distribution with known mean and variance
fontsize: 12pt
---

# R as a calculator: hands on experience

This short handout shows R being used as a calculator.  Follow the
examples through until you understand them, make sure you can
reproduce them in your own R session, and then try to complete the
exercises at the end.
       
The course is based around the _R computer language_.  This language
is a system that is designed to make analytics easy.  It is very
powerful but this lab is designed to get you familiar with R's most
basic functions: as a simple calculator.  We will cover more of R's
abilities in the remainder of the course.

This document assumes that you can log in to your computer, fire up R,
type commands into its command window, and read the output.  If you
have any difficulty, ask your TA.



# Try the simple stuff first

When using R (or any other language), it is important to try simple
things first:

```{r}
2+2
```

Notes on the above:

* This is the hardest part of the whole course.  You need to be
  able to fire up an R session, type stuff into it (in this case
  ``2+2''), and see the output.
* The ```R>```is known as the R prompt.  Text after that is
  typed by the user, using the keyboard.
* The response, ```4``` is given by R immediately after hitting "return".
  We will come to the   meaning of the ```[1]``` later.


Now, make sure that you can execute the 2+2 command in different ways.
Look at the ```Rmd``` file in the Rstudio app.  Change the 2+2 to 2+3
and execute the new command using the green arrow to the right, or the
pulldown "run" menu.  See how the answer changes from 4 to 5.

Now open the console and type "2+2" and "2+3" directly on the command
prompt.  Get a feel for how the interface works: the user types
something and when you hit return, R gives a response.

## Increase the complexity gradually

When you are confident with the addition examples, try something harder:

```{r}
(1+4)*(3-2)
```

Comments:
* Observe how brackets are used to evaluate the arithmetic in a specific order.
* An asterisk is used for multiplication

## Use common sense as an ongoing check

Always use common sense to anticipate the rough value of the answer
_before_ you execute a command in R.  For example:

```{r}
1000/7
```

Comments: 

* You might not know what $1000/7$ is exactly, but you will have
  an approximate idea.  You will know that $1000/10=100$, and maybe
  that $1000/8=112.5$, so you might guess that $1000/7$ is a little
  bit bigger than 112.5 but less than, say, $1000/5=200$.  So
  basically we can *estimate* the answer and verify that R is
  giving something close to the correct answer.
* When using R, stop and think.  Ask: _Is R doing what I think
  it is doing?_
* It is a good idea to verify _every single thing_ you do in R.


Another example:


```{r}
sqrt(60)
```

In R, ```sqrt()``` is the square root operator.  Again we don't know
what the square root of 60 is exactly.  But it's bigger than 7
(because $7^2=49$ which is less than 60) and it's smaller than 8
(because $8^2=64$ which is bigger than 60).  So I would have estimated
7.5 or thereabouts and this agrees roughly with R's answer.

Another example:


```{r}
pi^2
```

See how the caret ```^``` is being used for powers.  Well I know that
```pi``` (or $\pi$) is just over 3, about 3.14, and I also know that
$3^2=9$, so I'd estimate about 10.  So the exact answer agrees quite
closely with the exact answer.
It's important to sharpen your mental arithmetic to be able to
anticipate R's answer, because it's easy to get it wrong:

Going back to the first example, imagine you made a typo:

```{r}
1000/.7
```

(this is an easy mistake to make).  Now if you have an idea of the
rough size of any answer, you can catch this error straightaway (the
technical term is _recovery_).  But if you don't have an estimate,
errors like this can creep in, making any analysis meaningless.


## Another type of check


In arithmetic, we know that the order of addition makes no difference:
$a+b=b+a$.  So we can check this in R:


```{r}
23.223 + 10.123
10.123 + 23.223
```

So that checks out: both answers match.  The order of brackets makes
no difference either, so that $a+(b+c) = (a+b)+c$.  In mathematics,
this is called *associativity*. Check this:

```{r}
12.123 + (1.001-14.123)
(12.123 + 1.001)-14.123
```

Observe how the two lines have the same three numbers, but the
brackets are placed differently.  The third number is negative.  So
this checks out.


# Large numbers

Try some large numbers:

```{r}
10
100
1000
10000
100000
1000000
10000000
100000000
```

The first few are straightforward, but see how ````1000000``` (ie one
million) is displayed as ``1e+06```.  Try it!  This means that the
decimal point moves 6 places to the right.  Now try some small
numbers:

```{r}
1/10
1/100
1/1000
1/10000
1/100000
1/1000000
```

Again, the first few are straightforward, but very very small numbers
use the ```e``` notation.


## Help system

The R language comes with a huge amount of help.  Make sure you can
use it!

* In RStudio, the "help" pulldown menu gives all sorts of help, online
and local. Play with this.
* In plain R or RStudio, type ```help.start()``` at the commandline and
this opens a page of help.
* To get help on a particular R function, say the square root function
```sqrt()```, type ```?sqrt``` on the console.  This works in plain R
and RStudio.


You will be able to use all three of the above methods to get help,
even in the midsemester test or final exam. 

Of course, in a lab session you can use Google to find online
resources.  If you find any good resources, please tell your TA and
fellow students about them and then everyone can benefit.

