---
title: "MATPMDA lab, week 6, hypothesis testing"
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```


### Question

To test a null hypothesis $H_0\colon \mu=\mu_0$ for a population that
is $N(\mu, \sigma^2)$, it is usual to apply a transform so that we are
dealing with an observation that has a standard Gaussian, $N(0,1)$.
We know that

$$\overline{x}\sim N\left(\mu_0,\frac{\sigma}{\sqrt{n}}\right)$$

And we have been using R function `pnorm()` to calculate tail areas,
with something like

```
1 - pnorm(obs, mu_0, SEM)
```

where `obs` is the observation, `mu_0` is the hypothesised mean, and
`SEM` the standard error of the mean.  However, it is more convenient
to use a so-called "Z test".  If the null is true, we can say that
$\overline{x}-\mu_0\sim N\left(0,\frac{\sigma}{\sqrt{n}}\right)$ and
thus $\frac{\overline{x}-\mu_0}{\sigma/\sqrt{n}}\sim N(0,1)$.  We call
this quantity $Z$, and use it as a test statistic:

$$
Z = \frac{\overline{x}-\mu_0}{\sigma/\sqrt{n}}
$$

Again, if the null is true, we have $Z\sim N(0,1)$.

* Find the interval of $Z$ values that is the critical region (or
rejection region) for $\alpha=0.05$ (one-sided test).

```{r}
qnorm(0.05)
```

```{asis}
Therefore the critical region is the interval $(-\infty, -1.645)$.
```

* Find the critcial region for $\alpha=0.01$

```{r}
qnorm(0.01)
```


```{r}
alpha <- 0.05
qnorm(alpha/2, 0, 1)
qnorm(1-alpha/2, 0, 1)
```

```{asis}
Therefore the critical region is the interval $(-\infty, -1.96) \cup(1.96, \infty)$ or~$\{z\colon\left|z\right| > 1.96\}$
```

### Question

An army office is interested in the horizontal distance from the target
of a set of rifle shots, measured in centimetres.  It is known from
previous studies that the standard deviation of the horizontal
distance is 0.8 centimeters.

```{r,include=TRUE}
x <- c(-0.67, -0.03, 0.37, 1.87, 0.35, -0.47, 0.92, 0.95, 0.55, 0.16,
0.47, -0.44, 0.2, 0.11, 1.91, 1.06, -0.49, -0.93, 1.78, -0.31,
1.28, 0.26, 0.18, -0.26, 0.7, -0.37, -0.73, -1, 0.4, 1.03)
```

* Calculate the sample mean for this dataset.

```{r}
mean(x)
```

* Calculate the standard error for this dataset, using the known
  population standard deviation $\sigma=0.8$

```{r}
sigma <- 0.6
n <- length(x)
n
SEM <- sigma/sqrt(n)
SEM
```

* State a sensible null and alternative hypothesis to be used in this
  situation, in words and in mathematical symbols.

```{asis}
$H_0\colon \mu = 0$\\
$H_1\colon \mu \neq 0$

In words:

$H_0\colon$ the rifle shots have a mean horizontal distance of zero\\
$H_1\colon$ the rifle shots do not have a mean horizontal distance of zero
```

* State the precise definition of $p$-value, and explain what "more
  extreme" means in this context.

```{asis}
The probability, if the null is true, of obtaining the observation or an observation more extreme
```

* Compute the test statistic, i.e. $Z = \frac{\bar{X}-\mu_0}{\sigma/\sqrt{n}}$

```{r}
mu0 <- 0
z <- (mean(x)-mu0)/SEM
z
```



```{asis}
* Using a significance level of 5\%, identify the critical region.

This is a two-tailed test, so the critical region is values of the
test statistic which are greater than $c$ or less than $-c$.  In this
case, we have a significance level of 5\% so $P(Z > c) = 0.025$ and
thus $P(|Z| > c) = P(Z>c)+P(Z<-c) = 0.05$.
```

```{r}
qnorm(0.975, 0, 1)

```{asis}

The Critical region is values of $Z$ which are greater than 1.96 or less
than -1.96, i.e $\{z\colon |z| > 1.96\}$

* Is the test statistic within the critical region? What conclusion can we draw?

```{asis}

Yes, it is in the upper tail.  Therefore, we should reject the null
hypothesis and conclude that the mean horizontal distance of the rifle
shots is not equal to 0.
```

* Assess the evidence provided by the data calculating a p-value.  NB:
  a typical exam question will ask you to state the precise definition
  of p-value and explain what "more extreme" means in this context.

```{asis}

The p-value is the probability, if the null is true, of obtaining the
observation or one more extreme.  In this context, "more extreme"
means "the sample mean of the dataset is further away from zero than
0.295".
```

```{r}
1-pnorm(z, 0, sd=1)
2*pnorm(abs(mean(x)), mu0, SEM, lower.tail=FALSE)
```

```{asis} 

The p-value less than the significance level of 5\%,
therefore the result of the hypothesis test is significant.  We may
reject the null hypothesis and infer that the rifle is indeed biased.
```

* (optional: confidence intervals not yet covered). Compute a 95\%
  confidence interval for the mean horizontal deviation of the rifle
  shots.

```{r}
qnorm(c(0.025, 0.975), mean(x), sd=SEM)
mean(x) + c(-1,1)*qnorm(0.975, 0, 1)*sigma/sqrt(n)
```

```{asis}

We are 95\% confident that the true mean $\mu$ is within the interval
(0.0803, 0.5097).  Notice that this interval does not contain
$\mu_0=0$.
```

## Question

The following dataset shows the level of lake Huron over 98 years:

```{r,include=TRUE}
huron <- 
c(580.92, 581.86, 580.97, 580.8, 579.79, 580.39, 580.42, 580.82, 
581.4, 581.32, 581.44, 581.68, 581.17, 580.53, 580.01, 579.91, 
579.14, 579.16, 579.55, 579.67, 578.44, 578.24, 579.1, 579.09, 
579.35, 578.82, 579.32, 579.01, 579, 579.8, 579.83, 579.72, 579.89, 
580.01, 579.37, 578.69, 580.33, 578.67, 579.55, 578.92, 578.09, 
579.37, 580.13, 580.14, 579.51, 579.24, 578.66, 578.86, 578.05, 
577.79, 576.75, 576.75, 577.82, 578.64, 580.58, 579.48, 577.38, 
576.9, 576.94, 576.24, 576.84, 576.85, 576.9, 577.79, 578.18, 
577.51, 577.23, 578.42, 579.61, 579.05, 579.26, 579.22, 579.38, 
579.1, 577.95, 578.12, 579.75, 580.85, 580.41, 579.96, 579.61, 
578.76, 578.18, 577.21, 577.13, 579.1, 578.25, 577.91, 576.89, 
575.96, 578.4, 577.68, 578.38, 578.52, 579.74, 579.31, 579.89, 
579.96)
```

Climate scientists are looking for evidence that global climate change has increased the level of the lake.
Historically, the level of the lake has been 579.3cm.

* State a sensible null and alternative hypothesis.

```{asis}
Let $\mu$ denote the mean level of the Lake Huron.

* $H_0: \mu = 579.3$
* $H_1: \mu > 579.3$

Note that the null hypothesis corresponds to the situation in which the lake level
has not changed from the historical value of  579.3cm.
```

* Calculate the sample mean of the dataset

```{r}
mean(huron)
```

* Calculate the standard error of the mean

```{r}
SEM <- sd(huron)/sqrt(length(huron))
SEM
```

*  State the precise definition of p-value and state what ``more
   extreme'' means in this context

```{asis}

The p-value is the probability, if the null is true, of obtaining the
observation or one more extreme.  In this context, ``more extreme''
means ``the sample mean of the dataset is higher than the observed
sample mean of 579''
```

* Calculate the p-value using a Z-test and interpret

```{r}
pnorm(mean(huron),mean=579.3,sd=sd(huron)/sqrt(length(huron)))
```

```{asis}

$p<0.05$ so there is evidence that the level has increased
```

* A historian suggests that the later observations are flawed.
  Perform a similar analysis on the first 70 observations.

```{r}
huron_older <- huron[1:70]
pnorm(mean(huron_older),mean=579.3,sd=sd(huron_older)/sqrt(length(huron_older)))
```

```{asis}
No evidence!
```

* Use R's `t.test()` instead and compare.

```{r}
t.test(huron, mu=579.3)
```


### Question

Consider the following dataset for five patients, A-E, showing
strength measurements before and after a training period.  We expect
the training to increase the strength of the patients.

```{r, include=TRUE}
before <- c(A=24.2, B=30.4, C=32.7, D=19.9, E=25.1)
after  <- c(A=25.5, B=31.2, C=32.9, D=20.4, E=25.6)
```

* State a sensible hypothesis
* Are we considering a one-sided or two-sided test?
* State the precise definition of $p$-value and explain what "more
  extreme" means in this context
*  perform a Student t-test and interpret.




### Question

Regardless of age, about 20\% of American adults participate in
fitness activities (the criterion was activity taking place at least
twice a week).  A random sample of 100 adults over 40 years old found
15 who exercised at least twice a week.  Is this evidence of a decline
in participation after age 40?  We will use $\alpha = 0.05$.

* State the null and alternative hypotheses.

```{asis}
$H_0\colon \pi=0.2$\\
$H_0\colon\pi <0.2$
```

* You will recall that a binomial distribution is approximately normal
  if $n$ is large, and that if $X\sim\operatorname{Bin}(n,p)$ then,
  approximately, $X$ is normal with $X\sim N(np,\sqrt{np(1-p)})$.
  Using this normal approximation for the binomial, compute the test
  statistic $Z$. Recall that $Z = \frac{\hat{p} -
  \pi_0}{\sqrt{\frac{\pi_0(1-\pi_0)}{n}}}$ where $\pi_0$ is the
  hypothesised probability of success [e.g. for a coin-tossing
  experiment we would have $\pi_0=1/2$].

```{r}
x <- 15
n <- 100
phat <- x/n
pi0 <- 0.2   # null
z <- (phat - pi0)/(sqrt(pi0*(1-pi0)/n))
z
```

```{asis}
Observe that the mean and variance are calculated using the  *true*
value of $p$, not the estimate $\hat{p}=r/n$.  What would you do if
your information about $p$ came purely from observing $r$ successes
out of $n$ trials?
```


```{asis}
* Compute the critical region for this test, stating whether and why
  you are using a one-sided or two-sided test.
```

```{r}
qnorm(0.05, 0, 1)
```

```{asis} 
The difference betwen the t test and the Z
test is the use of the estimated standard deviation $\hat{\sigma}$
rather than the true value $\sigma$ in the denominator.  By far the
hardest part of deriving the t distribution is proving that the sample
mean and the sample variance are independent
```

```{asis}
* Is the test statistic in the critical region? What conclusion can we draw?

No, the test statistic is not in the critical
region. Therefore we fail to reject the null hypothesis.  There is
insufficient evidence to conclude that the percentage of adults over
40 who participate in fitness activities twice a week is less than
20\%.  There is no evidence that the percentage of adults over 40 who
participate in fitness activities twice a week is less than 20\%.
```

* Use R to compute the p-value for the test statistic $z$.

```{r}
pnorm(z, 0, 1)
```

```{asis}
The p-value is greater than the significance level of 5\%, therefore we fail to reject $H_0$.
```

* Use R to compute the exact p-value using `dbinom()`.

```{r}
sum(dbinom(0:15,100, 0.2))
```

