---
title: "MATPMDA: lab handout for week 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The R programming language is a very powerful system for statistical
analysis that has many features and abilities.  We will start by
getting used to the R interface and use it to accomplish simple tasks.

Use the labs in conjunction with the course handbook, and other
resources such as videos and screencasts which are available on
Blackboard under the "resources" tab.


# Using R as a calculator

Fire up R or Rstudio (ask your tutor if you are not sure where to find
RStudio).  Type your commands at the command prompt.  R will respond
with the result of its computation.  RStudio includes other ways of
working which many people find more convenient.

When using R (or any other language), it is important to try simple
things first:
```{r,label=twoplustwo}
2+2
```

The basic idea with R is that we can type stuff into it (in this case
`2 + 2`) and then get some output (4).  The `R>` is known as the R
prompt.  Text after that is typed by the user, using the keyboard.
The response ("`4`") is given by R.  We will discover the meaning of
the `[1]` later.

Try some more examples and when you are confident with the addition
examples, try something harder:

```{r,label=morecompl}
(1+4)*(3-2)
```

Observe how round brackets are used to evaluate the arithmetic in a
specific order.  Try some more examples using the following symbols.


*  `+` Addition
*  `-` Subtraction
*  `*` Multiplication 
*  `/` Division
*  `^` Powers

In mathematics, operations are performed in a specific order.  This is
referred to as _Order of Operations_.  You may remember acronyms like
*BEMA* or *BEDMAS*. These stand for: Brackets, Exponents,
Multiplication/Division, Addition/Subtraction.  The operations should
be performed in this order.

## Use common sense as an ongoing check

Always use common sense to anticipate the rough value of the answer
_before_ you execute a command in R.  When using R, stop and think.
Ask , "_Is R doing what I think it is doing?_"  It is a good
idea to verify *every single thing* you do.


### Example 1

Suppose we have \$1000 and want to divide it between 7 people. How much will each person get?

```{r,dividebyseven}
1000/7
```

Does this answer make sense? How can we check?  Here is an example of
the type of logic we could use to check R's responses.  You might not
know what~$1000/7$ is exactly, but you will have an approximate idea.
You will know that $1000/10=100$, and maybe that $1000/8=112.5$, so
you might guess that $1000/7$ is a little bit bigger than 112.5 but
less than, say, $1000/5=200$.  So basically we can _estimate_ the
answer and verify that R is giving something close to the correct
answer. 

### Example 2

R has some inbuilt commands, like `sqrt()`, which is the square root
operator.  Suppose we want to know the square root of 60.  We can
compute this in R, but we can also estimate the answer. How could we
estimate it?  

```{r}
sqrt(60)
```

Here is an example of how we could estimate it: we do not know what
the square root of 60 is exactly.  But it is bigger than 7 (because
$7^2=49$ which is less than 60) and it's smaller than 8 (because
$8^2=64$ which is bigger).  We could guess that it will be about 7.5.
And our guess is not too far off.

### Example 3

R also has a command for ``special'' numbers like $\pi$.  The command
`pi` will give us the value of 3.141593\ldots.  Suppose we want to
know the value of $\pi^2$.  Since $\pi$ is just over 3, and $3^2=9$,
we could estimate that $\pi^2$ is about 10.  Check with R.

        
```{r}
pi
pi^2
```

### Example 4

Going back to the first example, imagine you made a typo (and this is an easy mistake to make):

```{r}
1000/.7
```

Does this answer make sense?  Will each person get \$1428.57?!  Now if
you have an idea of the rough size of any answer, you can catch this
error straight away (the technical term is _recovery_).  But if you do
not have an estimate, errors like this can creep in, making any
analysis meaningless.


# Verify verify verify

It is very easy to make mistakes in mathematical work.  This means
that it is important to check your work as you go along.  That way you
can find your mistakes and correct them before they get serious.

For example, the result of dividing \$1000 by 7 was \$142.86, to the
nearest cent.  We can check this by multiplying our value by 7 and
hope that we get to \$1000:

```{r}
142.86 * 7
```

Not exactly right, but very close.  The slight difference is caused by
our rounding the value to two decimal places.  Consider the
$\sqrt{60}$ calculation above.  The answer given by R was about `7.745`.
This value will have a square of 60 if everything is correct:

```{r}
7.745^2
```

Again, very close to the expected value. 


# Large numbers and small numbers

Remember that the caret ("`^`") gives powers.  Try the following:

```{r}
10
10^2
10^3
10^4
10^5
10^6
10^7
10^8
10^9
10^10
10^11
10^12
```


The first few are straightforward, but see how `10^6` (ie one million)
is displayed as `1e+06`.  This means that the decimal point moves 6
places to the right.

Try some of these:

```{r}
20.3^23
0.3^-65
```

and see if you can understand the output.  Small numbers work in a similar way:


```{r}
0.1^30
10^-30
```

# Variables

Suppose we want to define a variable `x` and use its value in the future:


```{r,usex}
x <- 4.5
x
x^2
1/x
```

In the above, `x` is a variable whose value is 4.5 (the symbol "`<-`"
is read "becomes", so the first line reads "x becomes 4.5").  See how
we can use the value of `x` to avoid retyping its value.  We can go on
to assign other variables.  A variable name has to begin with a letter
or a dot, and may contain underscores.  Variable names are
case-sensitive, so `x` differs from `X`.


# Vectors

A _vector_ is a sequence of numbers that behaves as a single unit.


```{r,vectorusage}
v <- c(1,5,4,1,1,2)
v
```

Object `v` thus holds six numbers and we can treat them together:

```{r}
v + 1000
v^2
1/v
```

See how each element of `v` is manipulated separately.  By far the
most common vector is a simple sequence of numbers, which can be
created using the colon:


```{r}
1:9
100:120
```


## Including Plots

It is easy to create plots using the `plot()` command:



```{r plotv, echo=TRUE}
v
plot(v)
```

We can combine ideas from the previous examples:

```{r}
x <- 1:30
plot(x,x^2)
```

# Data analysis

We will now explore the Motor Trend Cars Dataset discussed in
lectures.  This dataset is part of R so it is already installed.
```{r,usemtcars}
mtcars        #prints whole dataset
head(mtcars)  #prints first 6 rows
tail(mtcars)  #prints last 6 rows
```

You can find more information about the dataset by typing `?mtcars` at
the command line.  To access specific columns of the dataset you can use a variety of commands


```{r}
mtcars$mpg        #prints mpg column - note $ is proceeded by a column name
mtcars[,"mpg"]    #prints mpg column - note that " " are required
mtcars[,1]        #prints the first column, which in this case is mpg
```

We can now create a scatter plot to investigate the relationship
between miles per gallon and vehicle weight.  What do you observe?

```{r}
plot(mtcars[,"wt"], mtcars[,"mpg"], xlab="Weight (lb/1000)", ylab="Miles per gallon", pch=16)
```

(the above could be accomplished by typing `plot(mpg~wt,
data=mtcars)`).  The scatterplot shows that as vehicle weight
increases, miles per gallon decreases.  This indicates that there is a
negative relationship between vehicle weight and miles per gallon for
the cars in the mtcars dataset.

Try plotting other variables in the dataset against one another in the
same way.  You may find `boxplot()` useful.

# Frequency tables

Sometimes we might be interested in discrete variables, and ask
questions such as "How many manual and automatic cars are in the
dataset?".  We can construct a table to show the frequencies.

```{r}
table(mtcars$am)
```

Create a table to show the number of cars with  either 3, 4 or 5 gears:


```{r}
gears <- table(mtcars[,"gear"])
gears
```

Which number of gears is most common?  We can present this information
graphically using `barplot()`:

```{r}
barplot(gears)
```


Modify your code in the previous question to add labels on the $x$ and $y$ axes and a title.



# Histograms

As datasets get bigger, histograms provide a better choice for visualising numeric data.
We can create a histogram of the horsepower data:

```{r}
hist(mtcars[,"hp"])
```

It is possible to customise the histogram by adding a title, axis labels and axis scales:


```{r}
hist(mtcars[,"hp"],  xlab="gross horsepower", 
   main="Histogram of Horsepower for Vehicles\n 
   in the 1974 Motor Trend Magazine ", ylim=c(0, 12), xlim=c(0, 400))
```

Compare the number of cars in each ``bin'' in your histogram with the
stem and leaf plot above (scale=1). Do you notice anything odd?

\begin{solution}[3cm]


# Boxplots 

Boxplots are also very useful to displaying numeric data.


```{r}
boxplot(hp~gear, data=mtcars, main="Horsepower for Vehicles \n  
             in the 1974 Motor Trend Magazine (n=32)")
```



## Further work

Your task is to analys the trees dataset which is built in to R.  Type
`?trees` to find out more about the dataset.  Use R commands similar
to those above to answer the following questions:

* How many trees in the sample?
* What range of values do each of the variables take?
* Is there a relationship between Girth and Height?
* Is there are relationship between Girth and Volume?

R has many other in-built datasets.  To see a list of the datasets
type `library(help = "datasets")`, or `data()`.  For the remainder of
the lab, apply the techniques in this lab to a new data set.

