---
title: "MATPMDA lab, week 4, normal distribution"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```


### Question

 Let $Z$ be a standard normal random variable.

* Find $p(Z \leq 1.1)$ 

```{r}
pnorm(1.1,mean=0,sd=1)
pnorm(1.1)  # using defaults; see ?pnorm
```

* Check your answer using random sampling and visual graphics

```{r}
table(rnorm(1e6) <= 1.1)
table(rnorm(1e6) <= 1.1)
hist(rnorm(1e6),nclass=40)
abline(v=1.1,lwd=5)
```

```{asis}
Observe that the numerical results above differ slightly due to random variability
```

* Find $p(Z \leq 2.25)$

```{r}
pnorm(2.25,mean=0,sd=1)
pnorm(2.25)  # using defaults; see ?pnorm
table(rnorm(1e6) <= 2.25)
table(rnorm(1e6) <= 2.25)
```

* Find $p(Z > 0.6)$

```{r}
1-pnorm(0.6)
table(rnorm(1e6) > 0.6)
```

* Find $p(0.4 \leq Z\leq 1.3)$ and verify numerically

```{r}
pnorm(1.3) - pnorm(0.4)

## numerical verification slightly harder:

z <- rnorm(1e6)
table((z <= 1.3) & ( z >= 0.4))
table((z <= 1.3) & ( z >= 0.4))
```


### Question

Now suppose we change the mean and standard deviation from the default
values to $\mu = 11.1, \sigma=2.4$; so now random variable $X\sim N(11.1,2.4)$.
Calculate:

* $p(X<12)$

```{r}
pnorm(12,mean=11.1,sd=2.4)
pnorm(12,11.1,2.4)   # naming of arguments optional
table(rnorm(1e6,11.1,2.4) < 12)
table(rnorm(1e6,11.1,2.4) < 12)
```

* $p(10.1\leq X\leq 10.3)$

```{r}
pnorm(10.3,11.1,2.4)-pnorm(10.1,11.1,2.4)

z <- rnorm(1e6,11.1,2.4)
table((z<10.3) & (z>10.1))
table((z<10.3) & (z>10.1))
```

* Transform $X$ to a random variable with standard normal distribution
and perform the same calculations.
  
```{r}
Z <- (12-11.1)/2.4
pnorm(Z)

z1 <- (10.1-11)/2.4
z2 <- (10.3-11)/2.4
pnorm(z2)-pnorm(z1)

ran <- rnorm(1e6)
table((ran<z2) & (ran>z1))
table((ran<z2) & (ran>z1))
```

### Question

Let $Z \sim N(0, 1)$, a standard normal distribution.

* Find the value of $z$ such that $P(Z \leq z) = 0.95$. 

```{r}
qnorm(0.925)
```

* Find the value of $z$ such that $P(Z > z) = 0.025$

```{r}
qnorm(1-0.025, 0, 1)
```

* Find the value of $z$ such that $P(Z < z) = 0.9$; verify numerically

```{r}
qnorm(0.9)
table(rnorm(1e6) < qnorm(0.9))
```


### Question


A particular brand of chocolate bar has an advertised weight of 45g.
On average the weight is 45g, however there is some variation between
individual chocolate bars.  Suppose that the weight of the chocolate
bars is normally distributed. If 15\% of the time the chocolate bars
weigh more than 50g, what is the standard deviation of the weight of
the chocolate bars?

```{asis}

$Z = \frac{X - \mu}{\sigma}$

$Z=\mathrm{qnorm(0.85)}$ or about 1.036433

We have $1.036433 = \frac{50 - 45}{\sigma}$, so we
can say $\sigma = 5/1.036433 = 4.824$.
```
```{r}
(50-45)/qnorm(.85)

# verify

z <- rnorm(1e6,mean=45,sd=(50-45)/qnorm(.85))
table(z>50)  # should be about 85%
```

### Question

Cerebral blood flow (CBF) in the brains of healthy people is normally
distributed with a mean of 74 and a variance of 256, as measured by a
particular type of medical machine.  Note that this means the standard
deviation is $\sqrt{256}=16$.

* What proportion of healthy people will have CBF readings between 60 and 80?

```{r}
mu <- 74
sd <- sqrt(256)
pnorm(80, mu, sd) - pnorm(60, mu, sd)
```

* What proportion of healthy people will have CBF readings above 100?

```{r}
1-pnorm(100, mu, sd) 
```

* If a person has a CBF reading below 40, he or she is classified as
at risk for a stroke.  What proportion of healthy people will be
mistakenly  diagnosed as ``at-risk''?

```{r}
pnorm(40, mu, sd) 
```


# Question

The following dataset shows length in cm of petals of a particular
iris species.  Give a 95\% confidence interval for the mean length.

```
(6.3, 5.8, 7.1, 6.3, 6.5, 7.6, 4.9, 7.3, 6.7, 7.2,
6.5, 6.4, 6.8, 5.7, 5.8, 6.4, 6.5, 7.7, 7.7, 6,
6.9, 5.6, 7.7, 6.3, 6.7, 7.2, 6.2, 6.1, 6.4, 7.2,
7.4, 7.9, 6.4, 6.3, 6.1, 7.7, 6.3, 6.4, 6, 6.9,
6.7, 6.9, 5.8, 6.8, 6.7, 6.7, 6.3, 6.5, 6.2, 5.9)
```

```{r}
x <- c(
  6.3, 5.8, 7.1, 6.3, 6.5, 7.6, 4.9, 7.3, 6.7, 7.2,
  6.5, 6.4, 6.8, 5.7, 5.8, 6.4, 6.5, 7.7, 7.7, 6,
  6.9, 5.6, 7.7, 6.3, 6.7, 7.2, 6.2, 6.1, 6.4, 7.2,
  7.4, 7.9, 6.4, 6.3, 6.1, 7.7, 6.3, 6.4, 6, 6.9,
  6.7, 6.9, 5.8, 6.8, 6.7, 6.7, 6.3, 6.5, 6.2, 5.9)
```

```{r}
qnorm(c(0.025,0.975),mean(x),sd(x)/sqrt(length(x)))
```

Question

If we have observations from a Gaussian distribution with unknown mean and standard deviation equal to 3, how many observations do we need to reduce the standard error to 0.1?

```{r}
(3/0.1)^2
```
