---
title: "stat500 lab, week 9, regression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```

# remember to include a scattergraph with any regression analysis

# no scattergraph = no course credit



### Question

A computer scientist in Auckland has a large number of computers in
his laboratory and suspects that more failures occur when the
temperature outside is higher.

The dataset is as follows:

```{r,include=TRUE}
temp <- c(16.1, 18.2, 16.5, 14.1, 15.2, 17.2, 17.9, 17.5, 16.8, 14.3)
fail <- c(9.90, 9.21, 10.01, 8.22, 9.33, 8.89, 9.58, 10.97, 10.38, 8.96)
```

The first dataset is the temperature in Auckland in Centigrade, the
second is the percentage of computers that have a failure.

* Construct a scatterplot of the data and describe the relationship that you observe.

```{r}
plot(fail~temp)
```

* Using `lm()`, perform a linear regression and interpret

```{r}
summary(lm(fail~temp))
```

```{asis}
pvalue exceeds 5\%, no evidence
```


* Interpret your findings in a way that a computer scientist with
a flaky air conditioning plant, who is not a statistician, could
understand. 

```{asis}

The p-value for the regression model as a whole is 0.1317, which
exceeds the usual significance level of $0.05$.

Whilst there appears to be a moderate positive relationship between
temperature and the number of failures, there is no statistical
evidence for this relationship.  The pattern observed in the graph is
likely due to random variation.
```

### Question


The following R commands define two datasets:

```{r,include=TRUE}
length <- c(0.111, 6.061, 0.171, 0.228, 0.802, 0.039, 0.64, 0.238, 0.603, 0.337)
conc <- c(0.093, 0.102, 0.087, 0.085, 0.015, 0.107, 0.044, 0.08, 0.026, 0.073)
```

Here, `length` is the length of a beetle measured in centimeters, and
`conc` is the concentration of a certain hormone found in a tissue
sample.  The researchers are studying whether there is a relationship
between the size of the beetle and the concentration of hormone
present.  The beetles are a small species and are known to always be
under 1cm in length.

[helpful hint: commands such as `plot(conc~length)` and might be
helpful here; to remove a single observation use a command like
`plot(conc[-2]~length[-2])`, which will remove the second observation
(ie observation number 2)]

* Plot the data and describe the relationship you observe.

```{r}
plot(length, conc)
plot(length[-2],conc[-2])
```

```{asis}
In the first graph, there is a clear outlier. Further inspection
reveals that this value may be an error, since beetles are known to be
under 1cm in length and observation 2 is over 6cm.
```

```{r}
plot(length[-2],conc[-2])
```

```{asis}

There appears to be a strong negative relationship between beetle
length and the concentration of the horomone.
```

* Perform the appropriate regression analysis.  Report the p-value and
  interpret the output.

```{r}
conc_new = conc[-2]
length_new = length[-2]
plot(conc_new~length_new)
fit <- lm(conc_new~length_new)
abline(fit)
summary(fit)
```

```{asis}
The p-value (2.62e-06) is less than 5\% therefore
significant.   There is statistically significant
evidence that there is a relationship between beetle
length and the concentration of the hormone.
```

*  Write the regression equation and use it to predict the level of
   hormone for a beetle measuring 0.5cm in length.

```{asis}
The equation is:
$\widehat{\mathrm{Hormone}} = 0.109 -0.118 \mathrm{Length}$
```

```{r}
0.109 - 0.118*0.5
```

```{asis}
The predicted hormone concentration is 0.05 units.
```

* Interpret the slope of the regression model.

```{asis}
The slope is -0.118. Therefore for a 1cm increase in beetle length we predict a decrease in the hormone concentration of 0.118.

Given what is known about the beetles (i.e. their maximum length is
1cm), we shouldn't use our model to predict hormone concentration for
beetles above this length.
```

### Question

* Load the household savings data (on Blackboard, file `HouseholdSavingsSurvey.csv'):

```{r,include=TRUE}
d <- read.csv("HouseholdSavingsSurvey.csv", head=TRUE)
head(d)
```

* Policy makers are interested in whether there is a relationship
between age and income level.  Construct a scatter plot of these two
variables.

```{r}
plot(Income~Age, data=d)
```


* Describe the relationship between these two variables.
```{asis}
There is a positive relationship between these two variables, however
as age increase, there is a wider range of incomes.
```

* Compute the correlation coefficient $r$ using the command 

```{r}
cor(d$Income,d$Age)
```

* Perform linear regression to assess the impact of age on income.

```{r}
summary(lm(Income~Age, data=d))
```

* What is the regression equation?
```{asis}
Income = -52792.7 + 2638.1 Age
```

* Interpret the value of the intercept. Does this make sense?

```{asis}
The intercept is -52792.7, this is the income which corresponds to someone with age 0.  This does not make sense.  Extrapolating results outside the range of data is often not a good idea (in this case age is from about 20 - 50).
```


