---
title: "MATPMDA lab, week 3, binomial distribution"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```


This tutorial focusses on the binomial distribution.  This tutorial
builds on chapter 3 of the course manual.


### Question

Let $X\sim\operatorname{bin}(13,0.8)$.  This means that we have 13 independent
  Bernoulli trials, each with a probability of 0.8 of success.

Compute the following probabilities using the R function
`dbinom(k,n,p)` and the formula: $P(X=k) =
\binom{n}{k}p^k(1-p)^{n-k}$.  Note that $\binom{n}{k} =
\frac{n!}{k!(n-k)!}$.  Check that you get the same answer using both
methods.

* $P(X = 10)$
```{r}
factorial(13)/(factorial(10)*factorial(13-10))*0.8^10*(1-0.8)^(13-10)

#alternatively
choose(13,10)*0.8^10*0.2^3

# better:

dbinom(10,13,0.8)
```


```{asis}
It is often useful to define variables and use these variables in the
functions.
```

```{r}
n <- 16
p <- 0.8
x <- 13
dbinom(x, n, p)
```

* $P(X \leq 9)$

```{asis}
there are a number of techniques we can use here
```

```{r}

dbinom(0,13,0.8) + dbinom(1,13,0.8) + dbinom(2,13,0.8) +
dbinom(3,13,0.8) + dbinom(4,13,0.8) + dbinom(5,13,0.8) +
dbinom(6,13,0.8) + dbinom(7,13,0.8) + dbinom(8,13,0.8) +
dbinom(9,13,0.8)


sum(dbinom(0:9,13,0.8))

pbinom(9,13,0.8)
```
```{asis}
(see how all the different methods give the same result).
```

### Question

According to one study, The proportion of adults (18 years or older)
who admit to texting while driving is 47\%. Suppose you randomly
select ten adult drivers and ask if they text while driving (do you
think this percentage is realistic?); you may assume the drivers are
independent.  Say $X$ is the number of drivers in the sample who admit
to texting while driving.  If $X$ is modelled by a binomial random
variable, state $n$ and $p$.

```{asis}
$X \sim\operatorname{bin}(10,0.47)$.\\
```

* Compute the probability mass function for all values of $X$ using R.

```{r}
n <- 10
x <- 0:10
p <- 0.47
y <- dbinom(x, n, p)
names(y) <-  x
y

# check that probabilities sum to 1:
sum(y)
```

* Construct a probability histogram of $X$.

```{r}

plot(x,dbinom(x,n,p),type='h')
```

* Compute the expected value and standard deviation for $X$ and verify numerically

```{r}
n*p  # mean
sqrt(n*p*(1-p))  # standard deviation
mean(rbinom(1e6, n, p))  # numerical simulation of mean
sd(rbinom(1e6, n, p))    # numerical simulation of sd
```

### Question

We take a fair coin (that is, one with probability $p$ of landing
heads = 0.5) and toss it $n=1000$ times.  

* Plot the probability of observing a total of $r=400,401,\ldots, 600$
heads.

```{r}
x <- 400:600
plot(x,dbinom(x,1000,1/2))
```

* Repeat the simulation experiment in the previous question using
different values of $p$.  

```{r}
f <- function(p){dbinom(0:1000,1000,p)}
x <- 0:1000
plot(x,f(0.5),type="l")
points(x,f(0.6),type="l",col="red")
points(x,f(0.7),type="l",col="orange")
points(x,f(0.8),type="l",col="yellow")
points(x,f(0.14),type="l",col="green")
```

### Question

A test contains 10 questions. Each question is multi-choice with four
options.  Let $X$ be the number of questions that a student answers
correctly.  Suppose that the student has not studied, so their
strategy is to randomly guess the answers.  This means that the
probability of guessing the correct answer is $p=1/4$.

* What distribution can be used to model the number of correct answers?

```{asis}
Binomial distribution, with $n=10$ and  $p=0.25$.
```
* What is the probability that the student answers exactly 5 questions correctly?

```{r}
dbinom(5, 10, 0.25)
```

* The pass mark for this test is to score 6 or more correct answers.
  What is the probability that our student passes the exam?

```{r}
sum(dbinom(6:10,10,1/4))
pbinom(5,10,1/4,lower.tail=FALSE)
```

```{asis}
A little under 2\%
```


### Question

(harder) Look at the help page for `rbinom` and see that it documents four functions: `dbinom()`, `pbinom()`, `qbinom()` and `rbinom()`.  Function `pbinom()` is the _cumulative probability density function, which gives $P(X\leq x)$.  Thus:

```{r}
n <- 10
r <- 6
p <- 1/3
sum(dbinom(0:6,10,1/3))
pbinom(6,10,1/3)
```

Above we see the same answer from both methods.

* Assess the accuracy and speed of the two methods.
* (harder) Try and figure out what `qbinom()` does.  Hint: the first argument, named `p` in the documentation, is the probability returned by `pbinom()`.  Try and plot this function and explain any sudden jumps.
