---
title: "STAT500: lab handout for week 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Getting help in R

There are a lot of resources available within R and on the web to help you learn to program in R.

* Typing  `?functionname` opens a webpage about the named function, e.g. `?plot` or `?seq`.
* Using two question marks broadens the search; `??searchterm` or `help.search("searchterm")` open a webpage with matches to a named search term
* Simply typing your query into google can often help.

# Vectors

Recall that in R, a _vector_ is a sequence of numbers that can be
treated as a single unit.  This can be a very effective and
time-saving technique.

Vectors can be created in a number of ways.  Make sure you understand
the following by either typing them out on the R commandline, or
editing this file and processing it directly.

First, the colon operator:

```{r}
1:10   
10:1    # decreasing sequence
```

we can vary the first number to start the sequence with a different number:

```{r}
1000:1020
```

Now the `seq()` and `c()` functions.  Remember to type `?seq` to get
help on this function, which includes a large number of arguments.

```{r}
seq(0, 20, 2) #  sequence from 0 to 20 by 2
c(1, 2, 5, 6)  # direct data entry
seq(from=100,to=83,by=-3)
```

## Extraction and replacement

If we want to extract a particular element of a vector, we use square
brackets:

```{r}
x <- c(17,3:10,100,7:2)
x
x[3]         # element number 3
x[4:7]       # elements 4 through 7
x[c(5,3,6)]  # elements 5,3,6
```

This gives the fourth element of x through to the sixth.  Remember to
check that R in fact does this.

We can also assign a particular element:

```{r}
x
x[5] <- 1000
x
```

The assignment then changes the fifth element of x to 1000.  We can do more sophisticated replacement:


```{r}
(x <- 1:10)
y <- c(50,55)
x[7:8] <- y
x
```

# Negative indices

Negative indices mean to omit elements:


```{r}
x <- 1:30
x
x[-5]
```

See how the fifth element of x is omitted.    Negative indices can be used to exclude more than one element:

```{r}
x <- 1:100
y <- x[-(50:70)]
```

# Summary  statistics

R includes many summary statistics:

```{r}
x <- c(1:19,4:6,100:101)
x
mean(x)
length(x)
sum(x)
```

# Logical Operators

R can also be used to answer questions using logical operations.


```{r}
2>5   # is 2 greater than 5?
```

Of course not.  It is `FALSE`.  The technique also works with vectors:


```{r}
(1:10) > 4.4
```


See how we get a vector of logicals.  Each element of the result is
the comparison of the vector on the left hand side `1:10`
with the right hand side `4.4`.


# Functions in R

Earlier we met the `sqrt()` function for square roots.  R has a lot of
in-built functions that can be used to analyse data, some of which are:


* `mean(x)`
* `median(x)`
* `table(x)`
* `diff(range(x))`
* `sd(x)`
* `var(x)`
* `min(x)`
* `max(x)`
* `summary(x)`


To explore these commands we will use the \texttt{mtcars} dataset which we saw last week.

First, compute the mean, median, and standard deviation of the
variable miles per gallon. Recall that the column of a dataframe can
be extracted using  the dollar, for example `mtcars$mpg`.

```{r}
summary(mtcars$mpg)
mean(mtcars$mpg)
median(mtcars$mpg)
sd(mtcars$mpg)
```

But we can compute the same statistics for just the first 9 cars using the extraction methods:

```{r}
summary(mtcars$mpg[1:9])
mean(mtcars$mpg[1:9])
median(mtcars$mpg[1:9])
sd(mtcars$mpg[1:9])
```


# Further Exercises

* Create the sequence $1, 4, 9, \ldots, 81, 100$
* Create the sequence  $200, 175, 150, \ldots, 25, 0$
* Define a sequence of integers starting from 14 and ending at 19.
* Define a sequence of integers $2,4,6,\ldots,20$.
* Define a sequence of integers $14,16,18,\ldots,100$.
* Define a sequence of integers $1,3,5,7,9,\ldots,101$.
* Define a sequence of integers $1,2,3,4,\ldots,98, 99, 100,102,104,106,108,\ldots, 138, 140$.



Explore the dataset `Titanic`.  Were first class passengers more
likely to survive than third class passengers?  Were male passengers
more likely to survive than female passengers?  