---
title: "stat500 lab, week 10, ANOVA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```

### Question

Consider the following dataset:

```{r,include=TRUE}

dd  <- data.frame(
  HairColour = structure(c(3, 3, 3, 3, 3, 1, 1, 1, 1, 1, 4, 4, 4, 
                           4, 4, 2, 2, 2, 2, 2), 
                           .Label = c("DarkBlond", "DarkBrunette", 
                           "LightBlond", "LightBrunette"), class = "factor"),
  Pain = c(62, 60, 71, 55, 48, 63, 57, 52, 41, 43, 
           42, 50, 41, 37, 45, 32, 39, 51, 30, 35)
)
head(dd,3)
tail(dd,3)
```
This dataset shows pain sensitivity as a function of hair colour.

* plot the dataset


```{r}
boxplot(Pain ~ HairColour, data=dd, 
     main="Pain sensitivity as a function of hair colour")
```

* State a sensible null hypothesis


```{asis}

In the case of an ANOVA, the null hypothesis is
that the mean value for each group ($i=1,2,3,4$) is identical.
The hypotheses can be written:

$H_0\colon\mu_1=\mu_2=\mu_3=\mu_4$ 

$H_1 \colon\mbox{at least one mean differs from the others}$.
```

* Perform an ANOVA analysis and interpret

```{r}
summary(aov(Pain~HairColour, data=dd))
```

```{asis}
pvalue less than 5\%, significant.  There is strong evidence that pain threshhold depends on hair colour.
```

### Question

The `npk` dataset is built into R.  To access it, type `npk` on the command line:

```{r,include=TRUE}
head(npk)
```

* Plot the data by block; use `boxplot()`.

```{r}
plot(yield~block,data=npk)
```

* From the graph, say whether there appears to be any evidence
  that yield depends on block.

```{asis}
Doubtful dependence; note the large difference in ranges between the block.
```
```{r}
boxplot(rnorm(24)~rep(1:4,6))
```

* Use ANOVA to assess whether the yield is a function of the block, and interpret.

```{asis}
Let $\mu_i$ denote the mean yield of peas (in
pounds/plot) for block $i$, $i=1, \ldots, 6$.

The hypotheses can be written:

$H_0\colon\mu_1=\mu_2=\mu_3=\mu_4 = \mu_5 = \mu_6$ 

$H_1 \colon\mbox{at least one mean differs from the others}$.
```

```{r}
summary(aov(yield~block,data=npk))
```

```{asis}

The p-value exceeds the 5\% cut-off so the result is _not_
  significant.  There is no evidence to suggest that the yield
depends on the block.
```


### Question

The _OrchardSprays_ dataset is built into R and refers to
effectiveness of different insecticide treatments on orchard trees;
read the R help page.

* Plot the decrease as a function of treatment.

```{r}
plot(decrease~treatment,data=OrchardSprays)
```

* Say whether it seems that the treatments have different effectiveness

```{asis}
From the plot, there appears to be a clear dependence.
```

* State a null hypothesis
```{asis}
Let $\mu_i$ denote the mean decrease in the volume of the solution for
treatment $i$, $i=1, \ldots, 8$,
	
The hypotheses can be written:

$H_0\colon \mu_1=\mu_2=\mu_3=\mu_4 = \mu_5 = \mu_6 = \mu_7=\mu_8$ 

$H_1 \colon\mbox{at least one mean differs from the others}$.
```


*  use ANOVA to assess whether the treatment affects the behaviour of bees.  

```{r}
summary(aov(decrease~treatment,data=OrchardSprays))
```

* Interpret your findings

```{asis}
The p-value less than 5\% therefore significant.


There are statistically significant differences in the average
solutions decrease between the different orchard spray
treatments. This implies that different orchard sprays do have
different impacts on repelling honey bees.
```

### Question

Look at the _warpbreaks_ dataset, which is about wool loom
performance, and read the helppage by typing `?warpbreaks` on the
command line.

```{r,include=TRUE}
head(warpbreaks,3)
tail(warpbreaks,3)
```

* Plot breaks as a function of wool type

```{r}
plot(breaks~wool,data=warpbreaks)
```

* State a sensible null

```{asis}

Let $\mu_i$ denote the mean number of breakages per loom
for wool type $i$, $i=A, B$.

$H_0\colon \mu_A = \mu_B$ 

$H_1 \colon\mbox{at least one mean differs from the others}$.
```

* Assess the dataset using `aov()`

```{r}
summary(aov(breaks~wool,data=warpbreaks))
```

* Interpret your findings

```{asis}
p-value = 10.8\% thus not significant.  There is no evidence for any
dependence between type of wool and the number of breaks per loom.
```

