---
title: "stat500 lab, week 11, Fisher's exact test"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include =FALSE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```

### Question

A Medical researcher is investigating whether calcium supplements
increase bone density.  He takes a sample of 32 subjects, of whom 15
are given calcium and 19 a placebo (a "placebo" is a dummy pill that
has no medical effect).  The researcher tests the subjects after a
year, and 15 subjects are found to have higher bone density.  Of these
15 subjects, 12 were given calcium.  We may represent the data in R as
follows:



```{r,include=TRUE}
M <- matrix(c(12,3,5,14),2,2)
dimnames(M) <- list(increased=c(T,F),given=c("calcium","placebo"))
M
```

* State a sensible null hypothesis

```{asis}

The null hypothesis is that the probability of having increased
bone density is the same for the treatment and placebo group. 

Symbolically

$$\operatorname{Prob}(\mathrm{increase}|\mathrm{calcium}) = 
  \operatorname{Prob}(\mathrm{increase}|\mathrm{placebo})
  $$
```

* State the precise definition of pvalue.  State what "more extreme" means in this context.


```{asis}
pvalue is the probability, if the null is true, of obtaining the observation 
or an observation more extreme.

Given the marginal totals, ``more extreme'' means more than 12 patients increasing bone density with calcium
```

* Perform a Fisher test and interpret


```{r}
fisher.test(M,alternative="greater")
```

```{asis}
pvalue significant so we can infer that calcium increases bone density
```




* (harder) Suppose only 11 of the 15 given calcium exhibited an
  improvement.  Does your conclusion change?
  
```{r}
M <- matrix(c(11,4,6,13),2,2)
dimnames(M) <- list(increased=c(T,F),given=c("calcium","placebo"))
M
fisher.test(M,alternative="greater")
```

```{asis}
still significant; result unchanged.
```

### Question

An engineering competition for undergraduate students challenges
entrants to construct a bridge with given materials.  
A total of 30 students enter the competition, of whom 10 are from AUT
and 20 from UoA.  Each student's bridge is tested, and 13 bridges pass
the test and 17 fail.  Of the 10 AUT students, 7 of them build bridges
that pass the test.  We may represent the data in R as follows:


```{r,include=TRUE}
M <- matrix(c(12,3,5,14),2,2)
dimnames(M) <- list(increased=c(T,F),given=c("calcium","placebo"))
M
```

* State a sensible null hypothesis

```{asis}

The null hypothesis is that the probability of having increased
bone density is the same for the treatment and placebo group. 

Symbolically

$$\operatorname{Prob}(\mathrm{increase}|\mathrm{calcium}) = 
  \operatorname{Prob}(\mathrm{increase}|\mathrm{placebo})
  $$
```

* State the precise definition of pvalue.  State what "more extreme" means in this context.


```{asis}
pvalue is the probability, if the null is true, of obtaining the observation 
or an observation more extreme.

Given the marginal totals, ``more extreme'' means more than 12 patients increasing bone density with calcium
```

* Perform a Fisher test and interpret


```{r}
fisher.test(M,alternative="greater")
```

```{asis}
pvalue significant so we can infer that calcium increases bone density
```




* (harder) Suppose only 11 of the 15 given calcium exhibited an
  improvement.  Does your conclusion change?
  
```{r}
M <- matrix(c(11,4,6,13),2,2)
dimnames(M) <- list(increased=c(T,F),given=c("calcium","placebo"))
M
fisher.test(M,alternative="greater")
```

```{asis}
still significant; result unchanged.
```

