---
title: "stat500 lab, week 11, Fisher's exact test"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```

### Question

A Medical researcher is investigating whether calcium supplements
increase bone density.  He takes a sample of 32 subjects, of whom 15
are given calcium and 19 a placebo (a "placebo" is a dummy pill that
has no medical effect).  The researcher tests the subjects after a
year, and 15 subjects are found to have higher bone density.  Of these
15 subjects, 12 were given calcium.  We may represent the data in R as
follows:



```{r,include=TRUE}
M <- matrix(c(12,3,5,14),2,2)
dimnames(M) <- list(increased=c(T,F),given=c("calcium","placebo"))
M
```

* State a sensible null hypothesis

```{asis}

The null hypothesis is that the probability of having increased
bone density is the same for the treatment and placebo group. 

Symbolically

$$\operatorname{Prob}(\mathrm{increase}|\mathrm{calcium}) = 
  \operatorname{Prob}(\mathrm{increase}|\mathrm{placebo})
  $$
```

* State the precise definition of pvalue.  State what "more extreme" means in this context.

```{asis}
pvalue is the probability, if the null is true, of obtaining the observation 
or an observation more extreme.

Given the marginal totals, ``more extreme'' means more than 12 patients increasing bone density with calcium
```

* One-sided or two?  Justify

```{asis}
One-sided; we expect the calcium to increase bone density
```

* Perform a Fisher test and interpret


```{r}
fisher.test(M,alternative="greater")
```

```{asis}
pvalue significant so we can infer that calcium increases bone density
```

* (harder) Suppose only 11 of the 15 given calcium exhibited an
  improvement.  Does your conclusion change?
  
```{r}
M <- matrix(c(11,4,6,13),2,2)
dimnames(M) <- list(increased=c(T,F),given=c("calcium","placebo"))
M
fisher.test(M,alternative="greater")
```

```{asis}
still significant; result unchanged.
```

### Question

An engineering competition for undergraduate students challenges
entrants to construct a bridge with given materials.  
A total of 30 students enter the competition, of whom 10 are from AUT
and 20 from UoA.  Each student's bridge is tested, and 13 bridges pass
the test and 17 fail.  Of the 10 AUT students, 7 of them build bridges
that pass the test.  We may represent the data in R as follows:


```{r,include=TRUE}
M <- matrix(c(7,6,3,14),2,2)
dimnames(M) <- list(uni=c("AUT","UoA"),test=c("passes","fails"))
M
```

* State a sensible null hypothesis

```{asis}

The null hypothesis is that the probability of having increased
bone density is the same for the treatment and placebo group. 

Symbolically

$$\operatorname{Prob}(\mathrm{increase}|\mathrm{calcium}) = 
  \operatorname{Prob}(\mathrm{increase}|\mathrm{placebo})
  $$
```

* State the precise definition of pvalue.  State what "more extreme" means in this context.

```{asis}
pvalue is the probability, if the null is true, of obtaining the observation 
or an observation more extreme.

Given the marginal totals, ``more extreme'' means a
smaller probability than observed, if the null is true.
```

* One-sided or two?  Justify
```{asis}

Sidedness unclear in this case.  We might argue that
we seek a difference between the universities, in which case
we would use a two-sided test.  Or we might argue that it is self-evident 
that AUT will have better students than UoA.

However, it is *not* acceptable to choose sidedness on the basis
of the observations.
```


* Perform a Fisher test and interpret


```{r}
fisher.test(M)  # two-sided by default
```

```{asis}
pvalue not significant so we can make no conclusion.
```

* (harder) Suppose that one of the AUT students whose bridge failed
the test is allowed to submit another bridge which passes (perhaps the
first test was unfair in some way).  Do your conclusions change?
  
```{r}
M <- matrix(c(8,6,2,14),2,2)
dimnames(M) <- list(uni=c("AUT","UoA"),test=c("passes","fails"))
M
```

```{asis}
significant; AUT is better.
```

### Question

A certain lady claims to be able to detect whether milk is added first
or last to her tea.  She is given 8 cups of tea of which 4 have had
milk added first and 4 milk second.  Of the 4 with milk first, she
identifies 3 correctly, and of the 4 with milk added second, also
identifies 3 correctly.  We may represent this as follows:


```{r,include=TRUE}
M <- matrix(c(3,1,1,3),2,2)
dimnames(M) <- list(milk_first_actual=c(T,F),milk_first_guess=c(T,F))
M
```



* State a sensible null hypothesis

```{asis}

The lady guesses at random.
```

* State the precise definition of pvalue.  State what "more extreme" means in this context.

```{asis}
pvalue is the probability, if the null is true, of obtaining the observation 
or an observation more extreme.

Given the marginal totals, ``more extreme'' means fewer errors.
```

* One-sided or two?  Justify
```{asis}

One-sided, she aims to guess correctly more often.
```

* Perform a Fisher test and interpret


```{r}
fisher.test(M,alternative="greater")
```

```{asis}
pvalue not significant so we can make no conclusion.
```

* (harder) Suppose she got all 8 cups correct.  Would your conclusion change?
  
```{r}
M[] <- c(4,0,0,4)
M
fisher.test(M,alternative="greater")
```

```{asis}
significant; she is a better guesser than chance would suggest
```
