---
title: "stat500 lab, week 7, Student t-test"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```
### Question

Thirteen babies are born in a maternity ward. Their lengths in centimetres are


```{r,include=TRUE}
babies <- c(49, 50, 45, 51, 47, 49, 48, 54, 53, 55, 45, 50, 48)
```

Over long previous experience the mean length at birth at this
hospital has been 50cm. Does this sample suggest that this mean has
changed?  Use a t-test to assess the evidence, with a 5\% level of
significance.

* Specify a null hypothesis $H_0$ and the alternative hypothesis $H_1$. 

```{asis}
The null hypothesis would be that the mean length of babies born in
this maternity ward is 50\,cm exactly.

(The null hypothesis is that the effect is absent, no difference,
nothing going on.)
```

* state the precise definition of $p$-value and explain what "more extreme" means in this context


```{asis}
The p-value is the probability, if the null is true, of
obtaining the observation (in this case, the sample mean) or an
observation more extreme.  Here, "more extreme" means "further away
from the hypothesised population mean".
```

* Use `t.test()` to assess the claim.

```{r}
t.test(babies, mu=50, alternative="two.sided")
```

* Interpret the results
	
```{asis}
The p-value exceeds the significance level of 5\%, therefore we fail
to reject the null hypothesis.  We conclude that the data are
consistent with a mean length at birth of 50\,cm.  There is no
evidence for any difference. 
```
  
### Question

* Baz decides to train himself as a life-saver and as part of his
fitness campaign he has to race against another swimmer, Bob.  The
times for swimming 100m are as follows:

```{r,include=TRUE}
baz <- c(14.2, 14.4, 12.3, 14.3, 12.6, 13.9, 12.1, 11.9, 15.2, 12.3)
bob <- c(13.7, 14.9, 13.6, 17.5, 13.9, 13.5)
```

We seek to determine whether there is a significant difference between
the two swimmers.  You may assume that the t-test is appropriate.

* Is a one-sided or two-sided test appropriate?   Justify your answer
```{asis}
A two-sided test is appropriate because the situation is symmetric
between the swimmers.  We have no reason to expect Baz to be
faster or slower than Bob.
```

* State a sensible null hypotheses.

```{asis}
Let $\mu_1$ denote Baz's mean time and $\mu_2$ denote Bob's mean time.\\

$H_0\colon\mu_1=\mu_2$\\
$H_1: \mu_1 \neq \mu_2$

```

* Is a paired or unpaired test indicated?
```{asis}
Unpaired: there is no natural pairing between the datasets

Furthermore, the sample sizes are different, which means we cannot do a paired test.
```

* What is the mean time for the two datasets?
```{r}
mean(baz)
mean(bob)
```

* Perform a t-test, and indicate whether there is a significant
difference at the 5\% level between the two swimmers.

```{r}
t.test(baz,bob,alternative="two.sided")
```

* Interpret the result of the t-test in a way that Baz and Bob
(who are not statisticians) could understand.

```{asis}
The p-value exceeds 5\% therefore there is not enough evidence to establish a difference between the swimmers.
```

### Question

A physics student measures the speed of light in a laboratory.  His results are as follows:

```{r,include=TRUE}
x <- c(299792459.1, 299792457.87, 299792457.59, 299792459.6, 299792457.74, 
299792460.08, 299792459.27)
```


(Note: all figures in meters per second; if you type
`options(digits=14)` on the commandline, you will see more the
significant figure on the output).  The true value of the speed of
light is known to be 299792458 exactly.  The student's professor says
that the equipment generally gives readings that are too high.

* State a sensible null hypothesis for this situation and an alternative

```{asis}
Null: the  mean value of all the student's readings is the true
speed of light exactly, namely 299792458.

Let $\mu$ denote the mean value of the speed of light obtained by the equipment in the lab.

$H_0\colon\mu = 299792458$

$H_1\colon\mu > 299792458$

See how poorly the English language deals with this issue.  The null
refers to the _ensemble_, not the dataset.  Also observe how difficult
it is to precisely state exactly what ensemble you are dealing with.

```

* State the precise definition of p-value and state what the
observation is, and what ``more extreme'' means in this context

```{asis}
The pvalue is the probability, if the null is true, of obtaining
the observation or a value more extreme.  Here, the observation is
the sample mean of 299792458.75.  In this context, "more
extreme" means "the sample mean is higher than 299792458.75".
```

* Conduct a t test and interpret.

```{asis}
The professor said that the equipment tended to over-read so a one-sided test is needed.
```

```{r}
t.test(x, mu=299792458,alternative="greater")
```

```{asis}
The p-value is less than 5\% so the test is significant and we may
reject the null.  There is evidence that the equipment is
over-reading.
```
