---
title: "stat500 lab, week 8, confidence intervals"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```


### Question

A manufacturer produces matchboxes.  The weights of the boxes are
normally distributed with a standard deviation of 0.9g, but unknown
mean.  He samples 10 matchboxes and finds their mean weight to be
10.1g.

* Give a 95\% confidence interval for the mean box weight.

```{r}
SEM <- 0.9/sqrt(10)
qnorm(c(0.025,0.975),10.1,SEM)
```

* Give a 99.9\% confidence interval for the mean box weight.

```{r}
qnorm(c(0.0005,0.9995),10.1,SEM)
```

### Question


Sunglasses of a particular make are measured for the UVA
transmittance, with the following results

```{r,include=TRUE}
sunnies <- c(1.24, 1.29, 1.4, 1.51, 1.31, 1.17, 1.35, 1.27, 1.33, 1.19)
```

* Calculate the sample mean and standard deviation

```{r}
mean(sunnies)
sd(sunnies)
```

* Use `t.test()` to give a 95\% confidence interval for the UVA
  transmittance.

```{r}
t.test(sunnies)
t.test(sunnies)$conf.int
```

* Give a 99\% CI

```{r}
t.test(sunnies,conf.level=0.99)$conf.int
```

* The manufacturer claims that the transmittance is less than 1.25\%.
Does the data support this?
  
```{asis}
One-sided test.
```

```{r}
t.test(sunnies,mu=1.25,alternative="greater")
```

```{asis}
No reason to reject the null
```

### Question

You will recall Baz and Bob from week 7.  These guys swim 100m and
record their times in seconds:

```{r,include=TRUE}
baz <- c(14.2, 14.4, 12.3, 14.3, 12.6, 13.9, 12.1, 11.9, 15.2, 12.3)
bob <- c(13.7, 14.9, 13.6, 17.5, 13.9, 13.5)
```

* Give a 95\% confidence interval for the difference in times between Baz and Bob

```{r}
t.test(baz,bob)
```
