---
title: "stat500 lab, week 5, estimation"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE) # change to TRUE when rendering solutions; this changes
                                      # the default visibility (which is overridden by explicitly
                                      # setting  include=TRUE in chunk options)
```


### Question

In an electrolysis experiment each student in a class of 30 chemistry
students measured the amount of copper precipitated from a saturated
solution of copper sulphate over a 30-minute period.  The sample mean
for all students was equal to 0.145 mole.  It is known that the
population standard deviation is 0.0051 mole.

* Compute the standard error of the mean.

```{asis}
Suppose $X$ is the amount of copper sulphate. 

There are 30 students, $n=30$.

The sample mean from 30 students is $\bar{x} = 0.145$

The population standard deviation $\sigma = 0.0051$.

The standard error of the mean is $SE[\bar{X}]=
\frac{0.0051}{\sqrt{30}}$
```


```{r}
0.0051/sqrt(30)
```

* Verify your result numerically

```{r}
sd(replicate(1e5,mean(rnorm(30,0.145,0.0051))))
```

* Find a 95\% confidence interval for the mean amount of copper
  precipitated from the solution of a 30-minute period 

```{r}
SEM <- 0.0051/sqrt(30)
c(0.145 + qnorm(0.975)*SEM, 0.145 - qnorm(0.975)*SEM)
```

### Question

An electric scale gives a reading equal to the true mass of an
object plus a random error that is normally distributed with mean 0
and standard deviation $\sigma = 0.1$ grams. Suppose that the results
of five successive weighings of the same object are as follows:

```{r,include=TRUE}
mass <- c(3.142, 3.163, 3.155, 3.150, 3.141)
```

* Determine a 95 percent confidence interval estimate of the true
mass (note that the standard deviation is known to be 0.1g).

```{r}
n <- 5 # number of observations
SEM <- 0.1/sqrt(n)
mean(mass) + c(-1,1)*qnorm(0.975)*0.1/sqrt(n)
```

* Determine a 99 percent confidence interval estimate of the true mass

```{r}
mean(mass) + c(-1,1)*qnorm(0.995)*0.1/sqrt(n)
```
```{asis}

The 99\% confidence interval is wider than the 95\% CI
```

### Question

The standard deviation of test scores on a certain achievement test is
11.3.  A random sample of 81 students had a sample mean score of 74.6.

* Find a 90 percent confidence interval estimate for the average score
of all students.

```{r}
74.6 + c(-1,1)*qnorm(0.95)*11.3/sqrt(81)
```

### Question

We wish to estimate $p$, the proportion of all newborn babies who are
male.  The gender of 10000 newborn babies was noted and  5061 were
male.

* Determine a 95 percent confidence interval for $p$.

```{r}
phat <- 5061/10000 # best estimate
phat + c(-1,1)*qnorm(0.975)*sqrt(phat*(1-phat)/n)
```

* Find the 99\% percent confidence interval 

```{r}
phat + c(-1,1)*qnorm(0.995)*sqrt(phat*(1-phat)/n)
```


### Question

From past experience it is known that the weights of salmon grown at a
commercial hatchery are normal with a mean that varies from season to
season but with a standard deviation that remains fixed at 0.18kg. 

* If we want to be 90 percent certain that our estimate of the mean
weight of a salmon is correct to within $\pm 0.05kg$, how large a
sample is needed? What if we want to be 99 percent certain?

```{asis}
\[\bar{X} \pm  z_{\alpha/2} \sigma/\sqrt{n}\]

If the margin of error is $MoE$ we might write

\[MoE = z_{\alpha/2} \sigma/\sqrt{n}\]

We can now solve for $n$ in R:
```
```{r}
sigma <- 0.18
z <- qnorm(0.95)
MoE <- 0.05
(z*sigma/MoE)^2
ceiling((z*sigma/MoE)^2)
```

* What would the answer be if we wanted to be 99\% certain?

```{r}
ceiling((qnorm(0.995)*sigma/MoE)^2)
```

### Question

A wine importer has the opportunity to purchase a large consignment of
a 1947 vintage wine.  Because of the wine's age, some of the bottles
may have turned to vinegar.  However, the only way to determine
whether a bottle is still good is to open it and drink some.  As a
result, the importer has arranged with the seller to randomly select
and open 20 bottles.  Suppose 7 of these bottles are spoiled. 

* Find the 95 percent confidence interval estimate of the proportion
of the entire consignment that is spoiled.

```{r}
n <-  20
phat = 7/n  # best estimate
z <-  qnorm(0.975)
phat + c(-1,1)*z*sqrt(phat*(1-phat)/n)
```


### Question

The dataset `Nile` is one of R's inbuilt datasets and shows historic
flow rates of the river Nile.  Type `?Nile` to get help on this
dataset.

* Estimate the mean annual flow of the river Nile and its standard deviation.

```{r}
mean(Nile)
sd(Nile)
```

* Construct a histogram of the flow and describe the distribution.

```{r}
hist(Nile)
```

```{asis}
looks reasonably normal
````

* Estimate the standard error of the mean

```{r}
n <- length(Nile) # number of observations
SEM <- sd(Nile)/sqrt(n)
SEM
```

* Compute a 95\% confidence interval for the mean flow of the river Nile.

```{r}
mean(Nile) + c(-1, 1)*qnorm(0.975)*sd(Nile)/sqrt(length(Nile))
```

NB: climate change will mean that this methodology is questionable as
the analysis assumes that the observations are not subject to
systematic drift.


